\documentclass[11pt]{article}

\usepackage[margin=0.82in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{tabularx}
\usepackage{array}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{float}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,fit,shapes.geometric}
\usepackage{hyperref}

\definecolor{navy}{HTML}{17365D}
\definecolor{blue}{HTML}{2F75B5}
\definecolor{lightblue}{HTML}{D9EAF7}
\definecolor{green}{HTML}{548235}
\definecolor{lightgreen}{HTML}{E2F0D9}
\definecolor{orange}{HTML}{C65911}
\definecolor{lightorange}{HTML}{FCE4D6}
\definecolor{red}{HTML}{C00000}
\definecolor{graybg}{HTML}{F3F5F7}
\definecolor{darkgray}{HTML}{404040}

\hypersetup{
  colorlinks=true,
  linkcolor=navy,
  urlcolor=blue,
  citecolor=navy,
  pdftitle={Prefix KV-Cache Policy Evolution: Simulator, Verifier, and Search Architecture}
}

\lstdefinestyle{python}{
  language=Python,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{navy}\bfseries,
  commentstyle=\color{green},
  stringstyle=\color{orange},
  backgroundcolor=\color{graybg},
  frame=single,
  rulecolor=\color{lightblue},
  breaklines=true,
  showstringspaces=false,
  columns=fullflexible,
  keepspaces=true,
  xleftmargin=0.5em,
  xrightmargin=0.5em
}

\lstdefinestyle{shell}{
  language=bash,
  basicstyle=\ttfamily\small,
  backgroundcolor=\color{graybg},
  frame=single,
  rulecolor=\color{lightblue},
  breaklines=true,
  showstringspaces=false,
  columns=fullflexible,
  keepspaces=true
}

\newcommand{\code}[1]{\texttt{\detokenize{#1}}}
\newcommand{\filepath}[1]{\nolinkurl{#1}}
\newcommand{\identifier}[1]{\nolinkurl{#1}}
\newcommand{\metric}[1]{\texttt{\detokenize{#1}}}
\newcommand{\important}[1]{\textbf{\color{navy}#1}}
\newcommand{\risk}[1]{\textbf{\color{red}#1}}
\newcommand{\good}[1]{\textbf{\color{green}#1}}

\setlist[itemize]{leftmargin=1.35em,itemsep=0.25em,topsep=0.35em}
\setlist[enumerate]{leftmargin=1.55em,itemsep=0.25em,topsep=0.35em}
\setlength{\emergencystretch}{2em}
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.55em}

\title{\textbf{\color{navy}Prefix KV-Cache Policy Evolution}\\
\large Simulator, Verifier, Workload, and Search Architecture}
\author{Technical Design Report}
\date{June 9, 2026}

\begin{document}

\maketitle

\begin{abstract}
This report documents the prefix KV-cache policy optimization system implemented.
The system searches for deployable,
compact admission and eviction scoring policies for an LLM prefix cache.  It combines
a deterministic prefix-tree simulator, a multi-workload verifier, deployable and
future-knowledge baselines, an AST-based complexity penalty, and an LLM-driven code
evolution loop.

It also positions the implementation against current KV-cache algorithms: production
exact-prefix designs, semantic-adaptive and workflow-aware replacement research,
distributed and tiered caches, non-prefix RAG reuse, and within-request KV reduction.
The comparison identifies which results are directly comparable and translates the
remaining state of the art into a prioritized simulator roadmap.

On the historical discovery-run \(8\)-token, \(48/96\)-block, agentic-aware verifier,
the promoted incumbent ranks second in the full reporting panel, behind the constrained
next-use oracle and ahead of all deployable credibility baselines.  It scores
\(77.113\), compared with \(97.074\) for the reporting-only oracle and \(70.362\) for
TinyLFU-LRU.  A subsequent production-oriented search on the operative \(16\)-token
verifier promoted a \(572\)-node policy scoring \(65.649\), ahead of TinyLFU-LRU at
\(63.548\).  It passes all seven checks in the held-out agentic surrogate gate and the
cyclic working-set tripwire with a \(0.2198<0.25\) token-hit gap.  It also scores
\(3.064\) on the hidden panel versus \(-2.789\) for TinyLFU-LRU and retains
substantially lower churn.
This was not a sub-\$5 one-shot discovery.  The production search and its required
simplification stage cost \$5.070 together; the retained lineage from the compact seed
cost \$21.981, and the documented pre-promotion research program cost at least
\$44.845 in recorded model API charges.
However, a post-promotion geometry sweep shows that this result is specific to
16-token blocks: the previous incumbent remains materially stronger at 8 tokens.
Its apparent charged-probe
regression on the discovery verifier is entirely due to source-complexity accounting:
raw probe behavior improves, agent churn falls, and cyclic behavior is unchanged.  The
new
\identifier{agentic_tool_workflows} training family and held-out structure probe expose
the remaining generalization weakness even when selection improves.  Detailed
historical search results are retained separately in
\filepath{docs/research_log.tex} rather than presented as current rankings.

The report emphasizes implementation details and trust boundaries: what the policy can
observe, what the simulator owns, what future information is quarantined for verifier
audits and oracle baselines, how each trial is scored, and how candidate source is
generated, isolated, evaluated, ranked, and persisted.  It also consolidates the
evidence-backed policy, verifier, and search principles discovered across the evolution
history, while keeping unresolved hypotheses explicitly separate.
\end{abstract}

\tableofcontents
\newpage

\section{System Purpose and Design Goals}

\subsection{Problem statement}

Prefix KV caching reuses prefill computation when a new request begins with a token
prefix that is already resident.  A serving system with finite cache capacity must make
two coupled decisions:

\begin{enumerate}
  \item \textbf{Admission:} after computing a missing prefix block, should that block be
  retained?
  \item \textbf{Eviction:} when capacity is exceeded, which legal resident block should
  be removed?
\end{enumerate}

The optimization target is not simply the highest average hit rate.  A usable policy
must generalize across request families, capacities, random seeds, phase changes,
multi-tenant traffic, scan pollution, long-running decodes, and priority classes.  It
must also remain implementable and understandable.  The verifier therefore rewards
reuse and service quality while penalizing churn, unfairness, wasted admissions,
policy-caused underfill, avoidable eviction decisions, and source complexity.

\subsection{Primary design goals}

\begin{itemize}
  \item \important{Determinism.} Every workload is reproducible from a family name,
  request count, block size, and random seed.
  \item \important{Deployability.} Candidate policies receive only information that
  could be maintained online.  Future-reuse fields are \code{None} for deployable
  candidates.
  \item \important{Simulator ownership.} Candidate code produces scores; it does not
  mutate cache state, choose illegal victims, or bypass prefix-tree invariants.
  \item \important{Robust ranking.} Validation spans multiple families, two capacities,
  and three seeds.  The score includes mean behavior, weakest workload-capacity
  behavior, and worst-seed behavior.
  \item \important{Interpretability.} The verifier emits fine-grained metrics and a
  top-level score breakdown rather than only a scalar.
  \item \important{Compact policies.} Complexity is charged with an unbounded concave
  AST-node penalty, with a bounded subsidy for canonical composable primitives, so dead
  hooks and unnecessary state remain costly.
  \item \important{Probe and hidden quarantine.} Recurrence-heavy structure probes are
  evaluated with normal runs but excluded from candidate selection; hidden families
  remain behind a separate final-reporting path.
\end{itemize}

\subsection{Non-goals}

The simulator is not a cycle-accurate GPU model.  It does not model allocator
fragmentation, per-layer byte footprints, host offload, multi-node routing, GPU kernel
contention, or real wall-clock latency.  Its purpose is to make policy quality
measurable under controlled, varied prefix-reuse patterns before hardware-specific
evaluation.

\section{Architecture Overview}

\subsection{Major components}

\begin{figure}[H]
\centering
\begin{tikzpicture}[
  node distance=0.75cm and 1.0cm,
  box/.style={draw=navy, rounded corners, fill=lightblue, align=center,
              minimum height=0.8cm, text width=3.0cm},
  verify/.style={draw=green, rounded corners, fill=lightgreen, align=center,
                 minimum height=0.8cm, text width=3.0cm},
  search/.style={draw=orange, rounded corners, fill=lightorange, align=center,
                 minimum height=0.8cm, text width=3.0cm},
  arrow/.style={-{Latex[length=2.5mm]}, thick, draw=darkgray}
]
  \node[search] (config) {YAML configuration\\and policy prompt};
  \node[search, right=of config] (levi) {Levi code evolution\\generation and mutation};
  \node[box, right=of levi] (candidate) {Candidate source\\policy factory};

  \node[verify, below=of candidate] (entry) {Isolated evaluator\\source loading, timeout};
  \node[verify, left=of entry] (evaluator) {Prefix KV-cache evaluator\\trial matrix and aggregation};
  \node[box, left=of evaluator] (sim) {Deterministic simulator\\prefix tree and cache state};

  \node[box, below=of sim] (workloads) {Synthetic workloads\\train, validation, hidden};
  \node[verify, right=of workloads] (metrics) {Trial metrics\\score and diagnostics};
  \node[verify, right=of metrics] (artifacts) {Reports and artifacts\\best program, JSON, Markdown};

  \draw[arrow] (config) -- (levi);
  \draw[arrow] (levi) -- (candidate);
  \draw[arrow] (candidate) -- (entry);
  \draw[arrow] (entry) -- (evaluator);
  \draw[arrow] (evaluator) -- (sim);
  \draw[arrow] (workloads) -- (sim);
  \draw[arrow] (sim) -- (metrics);
  \draw[arrow] (metrics) -- (evaluator);
  \draw[arrow] (evaluator) -- (entry);
  \draw[arrow] (entry) -- (levi);
  \draw[arrow] (metrics) -- (artifacts);
\end{tikzpicture}
\caption{System-level data flow.  The candidate emits scores; the simulator owns all
state transitions.  The evaluator converts trial metrics into the scalar returned to
Levi while preserving detailed artifacts.}
\label{fig:architecture}
\end{figure}

\subsection{Relevant source files}

\begin{longtable}{>{\raggedright\arraybackslash}p{0.34\textwidth}
                  >{\raggedright\arraybackslash}p{0.60\textwidth}}
\toprule
\textbf{Path} & \textbf{Responsibility} \\
\midrule
\endhead
\filepath{src/prefix_cache_evolve/evaluators/prefix_kv_cache.py} &
Core data model, deterministic simulator, metrics, workloads, baselines, complexity
calculation, aggregation, and scoring. \\
\filepath{src/prefix_cache_evolve/problems/prefix_kv_cache/evaluator.py} &
Levi-facing isolated entry points for path, source, callable, and hidden evaluation. \\
\filepath{src/prefix_cache_evolve/problems/prefix_kv_cache/runner.py} &
CLI, evolution workflow construction, artifact persistence, baseline reports, hidden
reports, and plots. \\
\filepath{configs/prefix_kv_cache.yaml} &
Evolution prompt, model and pipeline configuration, workload panel, and score weights. \\
\filepath{src/prefix_cache_evolve/workflow/execution.py} &
Adapter between Levi's \code{evolve_code} API and the repository's evaluator result
format. \\
\filepath{src/prefix_cache_evolve/workflow/config.py} &
Pydantic-validated YAML loading and conversion into arguments forwarded to Levi. \\
\filepath{src/prefix_cache_evolve/problems/prefix_kv_cache/incumbents/registry.json} &
Current role assignments for immutable production and discovery incumbent bundles. \\
\filepath{src/prefix_cache_evolve/problems/prefix_kv_cache/incumbents/production_16tok_20260609/policy.py} &
Exact promoted deployable source discovered by the production-oriented search. \\
\filepath{src/prefix_cache_evolve/problems/prefix_kv_cache/incumbents/discovery_8tok_20260608/policy.py} &
Exact retained discovery source used for the reproducible \(77.113\)
hardened-accounting result, originally reported as \(77.230\). \\
\filepath{src/prefix_cache_evolve/problems/prefix_kv_cache/incumbents/historical_compact_20260607/policy.py} &
Prior compact incumbent retained as a small alternate evolution seed. \\
\filepath{src/prefix_cache_evolve/tools/cli.py} &
Consolidated analysis, ablation, and deterministic tuning command tree. \\
\filepath{tests/test_prefix_kv_cache_evaluator.py} &
Functional PyTest coverage for policies, simulator invariants, metrics, workloads,
ranking behavior, reports, and evolution integration. \\
\bottomrule
\end{longtable}

\section{Candidate Contract and Trust Boundary}

\subsection{Required interface}

A candidate module must expose a compatible \code{build_candidate} or
\code{candidate_factory} factory.  The returned policy implements two scoring methods
and exactly three lifecycle callbacks:

\begin{lstlisting}[style=python]
def on_request_start(request: RequestInfo, now: int) -> None: ...
def on_cache_hit(block: PrefixBlockInfo, request: RequestInfo, now: int) -> None: ...
def on_cache_miss(block: PrefixBlockInfo, request: RequestInfo, now: int) -> None: ...

def score_admission(block: PrefixBlockInfo, now: int) -> float: ...
def score_eviction(block: PrefixBlockInfo, now: int) -> float: ...
\end{lstlisting}

Admission occurs only when \code{score_admission > 0}.  During eviction, the simulator
calls \code{score_eviction} for every legal victim and evicts the block with the
\emph{highest} score.  Therefore a high eviction score means ``evict sooner.''

\subsection{Candidate-visible request metadata}

\begin{tabularx}{\textwidth}{>{\ttfamily}p{0.31\textwidth} X}
\toprule
\normalfont\textbf{Field} & \textbf{Meaning} \\
\midrule
request\_id & Opaque per-trial request identifier.  It does not expose the sequential
generator index. \\
tenant\_id & Tenant namespace.  Prefix hashes include tenant identity, so cross-tenant
sharing is intentionally disabled. \\
session\_id & Request-only session metadata.  Blocks do not expose a session ID. \\
prompt\_length & Prompt length in tokens. \\
priority & Non-negative QoS signal.  It is not guaranteed to correlate with reuse. \\
request\_type & Normalized to the constant \code{request} before every candidate
callback.  Descriptive synthetic labels remain simulator-internal. \\
prompt\_tokens & Always replaced with an empty tuple before candidate callbacks,
including custom and trace-replay streams, preventing content fingerprinting. \\
predicted\_output\_length & Optional online estimate.  The true output length remains
simulator-internal. \\
recent\_admission\_pressure & Bounded 32-request online mean indicating how often prior
requests reached cache capacity.  It contains no workload-family label. \\
recent\_miss\_rate & Bounded 32-request online mean of prior per-request token miss
rates. \\
\bottomrule
\end{tabularx}

\subsection{Candidate-visible block metadata}

\begin{longtable}{>{\ttfamily}p{0.29\textwidth}
                  >{\raggedright\arraybackslash}p{0.65\textwidth}}
\toprule
\normalfont\textbf{Field} & \textbf{Meaning} \\
\midrule
\endhead
block\_id, prefix\_hash & Stable block key.  Candidate state should be keyed by one of
these values, never by Python object identity. \\
parent\_hash & Parent prefix block, or \code{None} for a root block. \\
depth & One-indexed block depth in the request prefix. \\
start\_token, end\_token, token\_count & Position and size of the block.  The final
block can be partial. \\
tenant\_id & Owning tenant. \\
created\_at, last\_accessed\_at & Logical arrival steps. \\
prev\_last\_accessed\_at & Previous observed prompt-chain access step, or \code{None}
before a prior access exists.  Online-computable from past and current arrivals. \\
last\_access\_gap & Logical-step distance from the current access to the previous
observed access, or \code{None} on first access. \\
access\_gap\_mean, access\_gap\_var & Fixed-state exponentially weighted first and
second-moment recurrence summary.  Both remain \code{None} until two gaps exist. \\
hit\_count & Historical hit count over the block's lifetime. \\
descendant\_count & Number of known descendant prefixes materialized so far. \\
active\_ref\_count & Number of active requests pinning the block. \\
subtree\_hit\_rate & Online hit rate aggregated over the block and all known
descendants. \\
subtree\_active\_ref\_count & Online active-reference total over the block and all known
descendants. \\
estimated\_recompute\_cost & Prefix-depth-sensitive deterministic recomputation cost. \\
\identifier{estimated_future_reuse}, \identifier{estimated_next_reuse_distance} &
\code{None} for deployable candidates; populated only for explicitly reporting-only
future-knowledge baselines. \\
\bottomrule
\end{longtable}

\subsection{Composable online primitives}

Candidates may optionally import \code{MultiTimescaleDecay}, \code{decay_vector}, and
\code{threshold_excess} from
\filepath{prefix_cache_evolve.problems.prefix_kv_cache.primitives}.
\code{MultiTimescaleDecay} maintains one to eight fixed-width exponential-decay
accumulators per key, bounds retained keys with deterministic least-recently-touched
eviction, and exposes \code{observe}, \code{observe_vector}, \code{values}, and
\code{combine}.  \code{observe} adds one amount to every channel, whereas
\code{observe_vector} adds a distinct amount per channel.  The latter lets a policy
represent bounded reuse, miss, or other online signals in one canonical state object
instead of hand-rolling several decay dictionaries.
\code{decay_vector} is the pure stateless counterpart.
\code{threshold_excess(value, threshold)} is a finite positive-hinge helper.  It was
added only after separate successful searches independently discovered persistent-
pressure and priority-aware pressure-excess terms.  It captures that repeated
functional form without constraining which online signal, threshold, coefficient, or
evidence interaction a policy may use.  These primitives expand the search space
without exposing future reuse or workload identity.

\subsection{What candidate code cannot control}

The simulator, not the policy, owns:

\begin{itemize}
  \item prefix hashing and prefix-tree structure;
  \item cache residency and capacity;
  \item active decode pins and release events;
  \item legal-victim filtering;
  \item admission rollback after pin-induced capacity failure;
  \item hit, miss, recompute, latency, churn, fairness, and audit accounting;
  \item hidden workload selection and future-knowledge audits.
\end{itemize}

Candidate methods returning non-numeric, boolean, infinite, or NaN scores are invalid.
Exceptions in factories, callbacks, or score functions also produce structured invalid
results.

\section{Simulator Design}

\subsection{Request and block construction}

Each workload produces a tuple of internal \code{WorkloadRequest} objects.  A request
contains candidate-visible \code{RequestInfo}, the simulator-only true output length,
the internal prompt tokens, and an optional logical arrival step.

The prompt is divided into physical cache blocks of \code{block_size_tokens}.  The
operative production-oriented default is sixteen tokens.  Synthetic traffic is
generated once at a fixed eight-token granularity, so block-size comparisons replay
identical token streams rather than changing prompt lengths with cache granularity.  For
each block depth \(d\), the simulator hashes the cumulative prefix:

\begin{equation}
  h_d = \operatorname{BLAKE2b}_{64}
        \left(\operatorname{repr}\left(
        \text{tenant\_id}, (t_1,\ldots,t_{\min(dB,L)})
        \right)\right),
\end{equation}

where \(B\) is the block size and \(L\) is prompt length.  Including the tenant ID
prevents accidental cross-tenant reuse.  Hashing the cumulative prefix rather than the
individual chunk ensures that identical chunks at different positions are not treated
as reusable prefix blocks.

The simulator materializes a rooted prefix tree lazily.  Internal block state includes
parent and child links, known descendant counts, global hit history, active reference
count, residency, and per-residency admission-accounting fields.

\subsection{Prefix-contiguous matching}

A request can reuse only the largest root-contiguous resident prefix.  Matching stops at
the first non-resident block:

\begin{equation}
  m(r) = \max\left\{k:
  b_1,\ldots,b_k \text{ are resident}\right\}.
\end{equation}

The simulator also enforces prefix-contiguous admission.  A block cannot be admitted
unless its parent is resident.  If admission is rejected or forced to bypass at depth
\(d\), deeper blocks remain misses and are not considered for admission.  This models
the reachability constraints of a real prefix cache rather than a generic unordered
object cache.

\subsection{Logical time and decode pinning}

The \code{now} argument is a logical arrival step, not wall-clock time.  Arrival steps
are monotonic but may repeat to model microbursts.  Requests without explicit arrival
steps default to sequential indices.

After a request begins, every hit or newly admitted block on its prefix is pinned for:

\begin{equation}
  D(r) = \max\left(1,
    \left\lceil\frac{\text{true output length}(r)}
                         {\text{active tokens per step}}\right\rceil
  \right).
\end{equation}

The default active generation rate is 64 tokens per logical step.  Release events
decrement active references when their logical time arrives.  Active blocks are never
legal eviction victims.  The candidate may see a predicted output length, but the true
length used for pinning is not exposed.

\paragraph{Optional shared prefix/decode capacity.}
The operative verifier remains in \code{prefix\_only} mode so historical selection
scores remain comparable.  An opt-in \code{shared} robustness mode additionally grows
decode KV at the configured active generation rate and charges it against the same
block capacity:

\begin{equation}
  N_{\text{prefix-resident}}(t) + N_{\text{decode-resident}}(t) \le C.
\end{equation}

Decode blocks are active and non-evictable.  New decode demand first evicts legal
inactive prefix leaves using the evaluated policy; if no legal prefix victim remains,
the simulator records failed decode-block allocations.  Failed demand is not retried,
but later demand continues to be measured, so any nonzero failure rate identifies an
infeasible demand trace rather than an executable schedule.  This makes the mode useful
for robustness analysis while leaving scheduler actions such as request rejection,
preemption, swapping, and separate prefix/decode budgets outside the current policy
surface.

\subsection{Per-request state machine}

\begin{figure}[H]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[
  node distance=0.55cm and 0.75cm,
  state/.style={draw=navy, rounded corners, fill=lightblue, align=center,
                text width=3.15cm, minimum height=0.72cm},
  decision/.style={draw=orange, diamond, fill=lightorange, align=center,
                   aspect=2.0, text width=2.1cm},
  terminal/.style={draw=green, rounded corners, fill=lightgreen, align=center,
                   text width=3.15cm, minimum height=0.72cm},
  arrow/.style={-{Latex[length=2.3mm]}, thick, draw=darkgray}
]
  \node[state] (advance) {Advance logical time; release expired pins};
  \node[state, below=of advance] (materialize) {Materialize request prefix chain};
  \node[state, below=of materialize] (start) {Call\\{\ttfamily on\_request\_}\\
  {\ttfamily start}};
  \node[state, below=of start] (match) {Match root-contiguous resident prefix};
  \node[state, below=of match] (hits) {Update and pin hits; call\\\code{on_cache_hit}};
  \node[state, right=1.8cm of hits] (miss) {For each remaining block: charge miss and call
  \code{on_cache_miss}};
  \node[decision, below=of miss] (admitq) {Admission score positive?};
  \node[state, below left=0.75cm and 1.6cm of admitq] (bypass) {Record policy bypass;
  block deeper admission};
  \node[state, below right=0.75cm and 1.6cm of admitq] (tentative) {Tentatively admit and
  pin};
  \node[decision, below=of tentative] (over) {Over capacity?};
  \node[decision, below=of over] (victimq) {Inactive resident leaf exists?};
  \node[state, below left=0.75cm and 1.8cm of victimq] (evict) {Score legal leaves; evict
  highest};
  \node[state, below right=0.75cm and 1.8cm of victimq] (forced) {Rollback new block;
  forced bypass};
  \node[terminal, below=4.3cm of bypass] (record) {Record request latency, churn,
  occupancy, QoS, and structural metrics};

  \draw[arrow] (advance) -- (materialize);
  \draw[arrow] (materialize) -- (start);
  \draw[arrow] (start) -- (match);
  \draw[arrow] (match) -- (hits);
  \draw[arrow] (hits) -- (miss);
  \draw[arrow] (miss) -- (admitq);
  \draw[arrow] (admitq) -- node[above left]{no} (bypass);
  \draw[arrow] (admitq) -- node[above right]{yes} (tentative);
  \draw[arrow] (tentative) -- (over);
  \draw[arrow] (over) -- node[right]{yes} (victimq);
  \draw[arrow] (over.east) -- ++(0.8,0) |- node[pos=0.25,right]{no} (record.east);
  \draw[arrow] (victimq) -- node[above left]{yes} (evict);
  \draw[arrow] (evict.east) -- ++(0.8,0) |- (over.west);
  \draw[arrow] (victimq) -- node[above right]{no} (forced);
  \draw[arrow] (bypass) -- (record);
  \draw[arrow] (forced) -- (record);
\end{tikzpicture}
}
\caption{Per-request simulator flow.  Miss callbacks continue for the remaining chain
even after admission becomes blocked, but deeper blocks are not scored for admission.}
\label{fig:request-flow}
\end{figure}

\subsection{Admission semantics}

For the first missed block whose parent path is still admissible:

\begin{enumerate}
  \item The simulator calls \code{score_admission}.
  \item A score at or below zero rejects the block, records deliberate policy bypass,
  and prevents deeper admission.
  \item A positive score tentatively makes the block resident and pins it.
  \item If capacity is exceeded, the simulator repeatedly evicts legal victims.
  \item If no legal victim exists because all leaves are pinned, the tentative block is
  rolled back and the rest of the request is forced to bypass.
\end{enumerate}

The newly admitted block is pinned before victim selection and therefore cannot
immediately evict itself.  An admission can evict one or more blocks and still fail if
the remaining over-capacity state has no legal victim.  The verifier counts those
evictions in churn and latency even though the new block is rolled back.

\subsection{Eviction legality}

Only inactive resident leaves are legal victims.  Interior nodes cannot be evicted
because doing so would make resident descendants unreachable.  Active leaves cannot be
evicted because an ongoing decode references them.

The legal set at time \(t\) is:

\begin{equation}
  \mathcal{V}_t =
  \left\{b:
  b.\text{resident}
  \land b.\text{resident\_children}=\varnothing
  \land b.\text{active\_ref\_count}=0
  \right\}.
\end{equation}

The simulator computes the policy score for every \(b\in\mathcal{V}_t\) and evicts:

\begin{equation}
  b^\star = \arg\max_{b\in\mathcal{V}_t}
             \operatorname{score\_eviction}(b,t).
\end{equation}

\subsection{Cost and latency proxy}

The estimated recomputation cost of a block is depth-sensitive:

\begin{equation}
  c(b) = b.\text{end\_token}\cdot c_{\text{prefill-token}},
\end{equation}

with \(c_{\text{prefill-token}}=1\) by default.  Summing this value across missed blocks
approximates the growing attention cost of recomputing progressively deeper prefixes.

For a request \(r\), the deterministic latency proxy is:

\begin{equation}
  \ell(r) =
  \sum_{b\in\text{misses}(r)} c(b)
  + 0.035\cdot\text{lookup blocks}(r)
  + 0.2\cdot\text{evictions}(r).
\end{equation}

A lookup probes all matched blocks and one failed block when the full prefix is not
resident.  The simulator reports p50, p95, and p99 latency proxies, high-priority tail
latency, recovery-phase latency, and p95 recompute cost.

\subsection{Useful and wasted admission accounting}
\label{sec:admission-accounting}

Admission quality is tracked per \emph{residency interval}, not per globally known
block.  Internal state contains:

\begin{lstlisting}[style=python]
admission_tracked: bool
resident_hit_count: int
\end{lstlisting}

When a block is tentatively made resident, these fields reset.  The interval becomes
tracked only after the block survives capacity enforcement.  Each later cache hit while
resident increments \code{resident_hit_count}.  The miss that caused the admission is
not a hit.

When a tracked block is evicted, or when the trial ends while it remains resident:

\begin{align}
  \text{useful admission} &\iff \text{resident hit count} > 0,\\
  \text{wasted admission} &\iff \text{resident hit count} = 0.
\end{align}

Rejected and forced-bypass attempts are neither useful nor wasted admissions because
they never became successful admissions.  If a block is admitted, evicted, and later
admitted again, the two intervals are classified independently.

The binary rates are:

\begin{equation}
  A_{\text{useful}} =
  \frac{N_{\text{useful}}}{\max(1,N_{\text{admitted}})},\qquad
  A_{\text{wasted}} =
  \frac{N_{\text{wasted}}}{\max(1,N_{\text{admitted}})}.
\end{equation}

For nonzero admissions the rates sum to one.  The verifier also reports
\metric{evicted_without_hit_rate}, which narrows the waste metric to admissions evicted
without an intervening hit.

The binary classification is retained for interpretation, but the verifier now also
weights admission outcomes by block token count.  For successful residency intervals
\(i\), let \(s_i\) be the block's token count and \(h_i\) its resident hit count:

\begin{align}
  T_{\text{admitted}} &= \sum_i s_i,\\
  T_{\text{useful}} &= \sum_{i:h_i>0}s_i,\qquad
  T_{\text{wasted}} = \sum_{i:h_i=0}s_i,\\
  T_{\text{saved-by-admission}} &= \sum_i h_i s_i.
\end{align}

The token-weighted waste rate is
\(T_{\text{wasted}}/\max(1,T_{\text{admitted}})\).  It is the admission-waste
quantity used by the score, so wasting a full block costs more than wasting a partial
block.  Realized admission utility is reported as:

\begin{equation}
  U_{\text{admission-token}} =
  \frac{T_{\text{saved-by-admission}}}
       {\max(1,N_{\text{admitted}}B)},
\end{equation}

where \(B\) is the configured block size.  This is the average number of full-block
equivalent token reuses delivered per successful admitted cache slot.  It distinguishes
a one-hit full block from a one-hit partial block and credits repeated hits.  The score
adds a small concave \(\log(1+U_{\text{admission-token}})\) reward, which makes the
metric actionable without double-counting saved tokens linearly.

\subsection{Verifier-only future audits}

The simulator maintains two future-reuse trackers:

\begin{itemize}
  \item The \textbf{candidate-visible tracker} is disabled for deployable candidates and
  enabled only for reporting-only future-reuse baselines.
  \item The \textbf{audit tracker} is always enabled internally, but its information is
  never included in candidate-visible block metadata.
\end{itemize}

The audit tracker supports \emph{avoidable eviction} accounting.  At each eviction, the
verifier compares the selected victim's next-use distance with every other legal
victim.  The eviction is avoidable if another legal victim would be reused later:

\begin{equation}
  E_{\text{avoidable}} =
  \mathbf{1}\left[
  d_{\text{next}}(b^\star)
  < \max_{b\in\mathcal{V}_t\setminus\{b^\star\}}
      d_{\text{next}}(b)
  \right].
\end{equation}

This is a regret diagnostic, not an oracle instruction to the policy.  A
short-reuse avoidable eviction is additionally recorded when the selected victim is
reused within eight logical steps while an alternative is reused later than eight
steps.  The constrained next-use oracle has zero avoidable evictions by construction.

The same boundary prevents indirect workload reconstruction.  Workload generation uses
family-specific random seeds internally, while every candidate factory receives the
separate fixed \code{policy_seed} value.  Candidate-visible request IDs are opaque,
synthetic request-type labels are normalized to \code{request}, and prompt-token
content is scrubbed before every callback.  Registered baselines cannot be marked both
deployable and future-reuse-dependent.

This is an experimental trust boundary, not an adversarial Python security sandbox.
Candidate source executes in an isolated process, but arbitrary malicious Python could
attempt reflection or inspect local source and artifacts.  Reported searches therefore
also retain static source checks and source audit; a capability-isolated policy process
or restricted DSL would be required for security against deliberately evasive code.

\section{Workload Model}

\subsection{Trial matrix and split policy}

The discovery full panel used 96 requests per workload, seeds \((11,23,37)\), block
size eight, and capacities \((48,96)\).  The operative production-oriented panel uses
block size sixteen and capacities \((24,48)\), preserving the same
\((384,768)\)-token capacity tiers.  A separate robustness report replays identical
token streams with block sizes \(8\), \(16\), and \(32\).

All headline numbers use synthetic streams generated from committed source; no private
or external production trace contributes to them.  The exact historical evaluator
settings are retained in
\filepath{configs/prefix_kv_cache_discovery.yaml}.  The committed discovery workload
manifest records summaries and SHA-256 fingerprints for all 81 ordered synthetic
streams, plus verifier \identifier{1.0.0}, the generator-source hash, and CPython
version.  Its aggregate
access-pattern fingerprint starts \identifier{4607782d2315}; the manifest retains
the full value.
Production-trace replay remains a separate user-supplied path; trace reports record the
input-file hash, and replay reports also record the derived request-stream hash.

The earlier incumbent scores \(62.641\) at the primary 16-token size under verifier
\identifier{1.0.0}.  The promoted
production search result now scores \(65.649\), ahead of TinyLFU+LRU at \(63.548\),
with validation token hit \(0.596\) and churn \(163.9\) per 1,000 requests.  At the
24-block capacity its token hit rate is \(0.579\) versus \(0.539\) for TinyLFU+LRU;
at 48 blocks it is \(0.638\) versus \(0.627\).  The earlier 32-token result remains a
warning that coarse granularity substantially damages irregular-prefix reuse.

\begin{center}
\begin{tabular}{lrrrr}
\toprule
\textbf{Split} & \textbf{Families} & \textbf{Capacities} & \textbf{Seeds} &
\textbf{Trials} \\
\midrule
Train & 5 & 2 & 3 & 30 \\
Validation selection & 10 & 2 & 3 & 60 \\
Structure-generalization probe & 2 & 2 & 3 & 12 \\
Normal evaluation, including probe & 17 & 2 & 3 & 102 \\
Hidden final report & 10 & 2 & 3 & 60 \\
\bottomrule
\end{tabular}
\end{center}

Each family receives a deterministic seed offset of \(1000(i+1)\), where \(i\) is its
position within its split.  The main combined score uses validation trials only.
Training metrics are emitted for diagnosis and prompt context.  Probe metrics are
reported during normal evaluation but excluded from the combined score, success flag,
and invalidity surcharge.  Hidden metrics are quarantined behind a separate evaluator
entry point.

The \code{--quick} mode uses 36 requests and seed \((3,)\).  It is intentionally labeled
smoke-only because prior runs demonstrated that a single-seed slice can materially
misrank the field.

\subsection{Workload families}

\begin{longtable}{>{\raggedright\arraybackslash}p{0.24\textwidth}
                  >{\raggedright\arraybackslash}p{0.11\textwidth}
                  >{\raggedright\arraybackslash}p{0.57\textwidth}}
\toprule
\textbf{Family} & \textbf{Default split} & \textbf{Behavior and failure mode exposed} \\
\midrule
\endhead
\identifier{shared_system_prompt} & Train &
Repeated shared system instructions, a small recurring task set, and request-specific
tails.  Rewards preservation of shallow broadly shared roots. \\
\identifier{rag_template_reuse} & Train &
Shared RAG template, position-stable recurring chunks, and query tails.  Avoids
incorrectly crediting arbitrary repeated chunks outside prefix alignment. \\
\identifier{long_context_mixed} & Train &
Recurring documents with varying prefix depths and partial tails.  Stresses
depth-sensitive recomputation value. \\
\identifier{session_continuation_growth} & Train &
Interleaved conversations that grow by one turn per revisit.  Stresses paused deep
history preservation. \\
\identifier{agentic_tool_workflows} & Train &
Six interleaved tool-using workflows fork from shared checkpoints, resume paused
routes, append variable multi-step observations, and periodically replan after a
rollback.  Trains irregular agentic reuse without duplicating the held-out probe. \\
\midrule
\identifier{agent_trace_branching} & Probe &
Shared agent roots and schemas with accumulating tool history, branches, and retries.
Tests trunks, fanout, deep histories, and rollback-like retry behavior. \\
\identifier{phase_shift_prompts} & Validation &
Popular prompt family changes halfway through the run.  Exposes stale-history
overprotection. \\
\identifier{multi_tenant_skew} & Validation &
Uneven tenant volume with tenant-specific roots.  Supplies the score's explicit
max--min tenant fairness penalty. \\
\identifier{hotset_cold_scan} & Validation &
Warm hot set, one-off scan, then hot-set return.  Tests pollution resistance and
recovery. \\
\identifier{cyclic_working_set_pressure} & Probe &
Repeatedly traverses working sets near and beyond cache capacity, then expands the
cycle.  Exposes short-horizon and avoidable eviction regret. \\
\identifier{concurrent_long_generation} & Validation &
Batched arrivals and long outputs create overlapping pins and forced-bypass pressure. \\
\identifier{reasoning_burst} & Selectable robustness &
Three-request microbursts combine recurring prefixes, noisy predicted output lengths,
and heavy-tailed reasoning outputs. \\
\identifier{reasoning_burst_shifted} & Selectable robustness &
Four-request microbursts with longer reasoning outputs and a shifted branch traversal. \\
\identifier{stochastic_serving_mix} & Validation &
Interleaves chat, RAG, agent, long-generation, and one-off traffic under changing
regimes, bursts, and arrival gaps. \\
\identifier{rolling_template_versions} & Validation &
Stable version, canary overlap, majority rollout, and rollback.  Tests stale-version
retirement and active-version warmth. \\
\identifier{heavy_tailed_prefix_lengths} & Validation &
Recurring documents with Pareto-like prefix depths.  Tests expensive long-tail
contexts. \\
\identifier{priority_burst_recovery} & Validation &
Recurring high-priority traffic before, during, and after low-priority scan pollution.
Rewards appropriate priority protection and recovery. \\
\identifier{priority_one_off_noise} & Validation &
Reusable normal-priority traffic competes with mostly unique high-priority requests.
Prevents learning that priority automatically implies reuse. \\
\midrule
\identifier{adversarial_unique_prompts} & Hidden &
Mostly unique prompts.  Exposes admit-everything churn and dead admissions. \\
\identifier{cross_family_mixture} & Hidden &
Mixture of shared, phase-shifted, and unique requests. \\
\identifier{tenant_session_reentry} & Hidden &
Paused tenant sessions reenter after unrelated traffic with new one-off tails. \\
\identifier{stochastic_serving_mix_shifted} & Hidden &
More bursty, more one-off, and different regime weights. \\
\identifier{rolling_template_versions_shifted} & Hidden &
Rolls forward to a new version with a changed root. \\
\identifier{heavy_tailed_prefix_lengths_shifted} & Hidden &
Heavier tail, more documents, and deeper maximum prefix. \\
\identifier{priority_burst_recovery_shifted} & Hidden &
Higher priority range, larger hot set, deeper scans, and stronger burst pressure. \\
\identifier{cyclic_working_set_pressure_shifted} & Hidden &
Larger cycles and a different traversal stride. \\
\identifier{priority_one_off_noise_shifted} & Hidden &
Higher one-off priority burst ratio and deeper one-off prompts. \\
\identifier{tenant_phase_shift_cycles_shifted} & Hidden &
Longer tenant phases, repeated pollution/recovery cycles, and larger delayed reentry
gaps. \\
\bottomrule
\end{longtable}

\section{Verifier and Metrics}

\subsection{Metric taxonomy}

The verifier emits metrics at trial, split, workload, and capacity aggregation levels.
Important categories include:

\begin{itemize}
  \item \textbf{Reuse:} token and block hit rate, saved prefill tokens, depth-band hit
  rates, role-specific hit contributions.
  \item \textbf{Request service:} per-request p10 and p50 hit rates, worst-quarter and
  final-quarter token hit, quarter-to-quarter variance.
  \item \textbf{QoS:} priority-weighted hit rate, high- and low-priority hit rates,
  high-priority p10 request hit and tail latency.
  \item \textbf{Admission:} attempts, successful admissions, rejections, policy bypass,
  forced bypass, binary useful/wasted admissions, admitted/useful/wasted token mass,
  saved tokens per admitted slot, policy-caused underfill, and evictions without a hit.
  \item \textbf{Eviction regret:} reuse after eviction, short-distance reuse misses,
  reuse-distance percentiles, avoidable evictions, and avoidable short-reuse
  evictions.
  \item \textbf{Cost and pressure:} recompute tokens and cost, lookup probes, p50/p95/p99
  latency proxy, churn per 1000 requests, total/prefix/decode occupancy, decode-block
  demand, allocation failures, decode-pressure evictions, active-request peak, and
  maximum prefill cost.
  \item \textbf{Fairness:} tenant max--min hit-rate gap, tenant p10 hit rate, and Jain
  fairness.
  \item \textbf{Robustness aggregation:} worst trial, p10 across trials, and
  cross-trial standard deviation.
  \item \textbf{Validity and complexity:} invalid reason/fraction, timeout and memory
  failures, and scoring-function AST complexity.
\end{itemize}

\subsection{Workload-capacity base score}

For one workload-capacity group \(G\), containing three seed trials, define:

\begin{align}
  B(G) ={}&
  80\,\overline{H_{\text{tok}}}
  +60\,\overline{H_{\text{blk}}}
  +12\,\overline{H_{\text{req,p10}}}
  +12\,\overline{H_{\text{worst-quarter}}} \nonumber\\
  &+8\,\overline{H_{\text{high-priority}}}
  +\overline{\log(1+U_{\text{admission-token}})}
  -6\,\overline{A_{\text{wasted,tok}}}
  -8\,\overline{E_{\text{avoidable}}}
  -C_{\text{latency}}(G).
\end{align}

The priority term is included only for groups that contain positive-priority requests.
The automatic latency normalizer is scoped to the same workload-capacity group:

\begin{align}
  L_{\text{norm}}(G) &=
  \max_{i\in G}\text{max-prefill-cost}_i,\\
  C_{\text{latency}}(G) &=
  \min\left(40,\,
    35\frac{\overline{\text{p95-latency}}}{\max(1,L_{\text{norm}}(G))}
  \right).
\end{align}

This scoping prevents a long-context family from making latency penalties negligible
for all shorter workloads.

\subsection{Worst-seed blending}

For workload-capacity group \(G\), the verifier computes the group-average base score
and the minimum single-seed base score:

\begin{equation}
  S_G = 0.85\,B(G) + 0.15\min_{i\in G}B(\{i\}).
\end{equation}

This retains a strong mean signal while charging policies whose performance collapses
on one seed.

\subsection{Combined score}

For trial \(i\), policy-caused underfill multiplies deliberate policy bypass by unused
mean capacity:

\begin{equation}
  U_i =
  \text{policy-bypass-token-rate}_i
  \max\left(0,1-\frac{\text{mean-prefix-occupancy}_i}{\text{capacity}_i}\right).
\end{equation}

This distinguishes a selective policy that leaves reusable cache space idle from
natural underfill on short or low-reuse workloads where every miss is admitted.  Decode
occupancy is deliberately excluded so active generation cannot mask a prefix policy
that bypasses useful admissions.

Let \(\mathcal{G}\) be the 20 validation workload-capacity groups.  Let
\(\overline{\chi}\) be mean validation churn per 1000 requests, \(\overline{U}\) be
mean validation policy-caused underfill, \(F\) be mean tenant max--min gap on
\code{multi_tenant_skew}, and \(C\) be candidate AST complexity.  The combined score
is:

\begin{align}
  S_{\text{combined}} ={}&
  \frac{1}{|\mathcal{G}|}\sum_{G\in\mathcal{G}} S_G
  +0.5\min_{G\in\mathcal{G}}S_G \nonumber\\
  &-\min(25,0.015\,\overline{\chi})
  -\min(15,12\,\overline{U})
  -\min(30,80F)
  -0.065\,C^{0.75}.
\end{align}

The weakest-workload contribution can be negative.  This is intentional: catastrophic
behavior on one workload-capacity pair should reduce the total even if the mean is
strong.

\subsection{Complexity accounting}

Raw complexity counts Python AST nodes in every executable module-level subtree, not
only direct class and function definitions.  This includes uppercase constants,
condition wrappers, expression statements, and lambda bodies.  Imports, the module
docstring, a literal \code{\_\_all\_\_}, and the direct
\code{candidate\_factory = build\_candidate} interface alias are excluded.  Thin
factory wrappers named \code{build\_candidate}, \code{candidate\_factory}, or
\code{run\_demo} are otherwise ignored, but nested classes, functions, and lambdas
inside them are counted.  A syntax error receives complexity 10,000 before failing
loading.

The current configuration enables form-aware complexity.  Each statically recognized
call site using \code{MultiTimescaleDecay} or \code{decay_vector} from the canonical
primitive module receives a three-node credit.  The smaller
\code{threshold_excess} helper receives one node, because its hidden implementation is
only a positive hinge rather than bounded state machinery.  Total credit is capped at
25\% of raw implementation nodes.  The cap prevents primitive calls from making
arbitrary code free; repeated dead calls still increase effective complexity.  Setting
\code{form_aware_complexity: false} restores the legacy raw-node schedule.

The penalty \(0.065C^{0.75}\) is:

\begin{itemize}
  \item \textbf{unbounded}, so large dead implementations never become free;
  \item \textbf{concave}, so necessary implementation detail is not charged linearly;
  \item \textbf{behavior-independent}, allowing complexity to break near-ties between
  policies with similar outcomes.
  \item \textbf{form-aware but bounded}, mildly favoring reusable online primitives
  without suppressing the cost of custom control flow or state.
\end{itemize}

The penalty remains smooth and unbounded, but search also uses separate exploration
and promotion limits.  The general lane may evaluate proposals modestly above 650
effective nodes, but marks them exploration-only and directs their next mutation to
perform dedicated behavior-preserving simplification.  The eviction-specialist lane
exposes only a victim-ranking function and explores up to 1000 effective nodes so
richer eviction ideas can be measured.  Search ranks raw behavior before complexity,
preventing deletion alone from appearing as progress.  Final promotion retains 650 as
the deployable-source limit.  Thus the evaluation safeguards control pathological
growth, while the promotion limit controls deployability.

The promotion gate is paired with fail-closed source validation.  Candidate modules
may import only \code{math} and the documented
\code{prefix\_cache\_evolve.problems.prefix\_kv\_cache.primitives} symbols.
Executable top-level wrappers, nonliteral module state, decorators, dynamic execution
or compilation, reflective builtins, dunder access, broad exception handlers, and
future-knowledge fields are rejected before loading.  These restrictions prevent a
candidate from laundering implementation code through syntax that receives little or
no complexity charge, or from importing a repository baseline as a nominally tiny
wrapper.

\subsection{Invalid candidates, isolation, and memory}

Candidate evaluation normally runs in a forked worker, when nested forking is
available, with the operative YAML wall-clock timeout of 90 seconds and a
matching POSIX CPU limit.  Candidate Python allocations
are monitored with \code{tracemalloc}; the default peak limit is 64 MiB.  On Linux,
the worker also receives an \code{RLIMIT\_AS} ceiling equal to inherited virtual
memory plus fixed evaluator headroom and the configured candidate budget.

This boundary is resource-limited process isolation, not a complete security sandbox.
The child still inherits host filesystem, environment, and network access.  Untrusted
model-generated programs therefore require a separate OS or container sandbox with
the repository and verifier mounted read-only.

Any positive invalid fraction bypasses normal scoring:

\begin{equation}
  S_{\text{invalid}} =
  -1000 - 1 - 1000\,f_{\text{invalid}}.
\end{equation}

This keeps invalid programs below representative valid programs even when the valid
program is very complex.

\subsection{Baseline and oracle policies}

\begin{longtable}{>{\ttfamily}p{0.24\textwidth}
                  >{\raggedright\arraybackslash}p{0.16\textwidth}
                  >{\raggedright\arraybackslash}p{0.52\textwidth}}
\toprule
\normalfont\textbf{Policy} & \textbf{Class} & \textbf{Purpose} \\
\midrule
\endhead
\identifier{no_cache} & Deployable & Lower bound and bypass sanity check. \\
lru & Deployable & Recency baseline; evicts least recently used legal leaf. \\
\identifier{sglang_radix_attention} & Deployable & SGLang RadixAttention's default
radix-cache replacement policy; retains cache-page-aligned prefixes and recursively
evicts the least-recently-used unreferenced leaf.  Registered as a selectable
reference but excluded from default comparisons because it maps to leaf-LRU here. \\
\identifier{vllm_apc} & Deployable & Behavioral emulation of the core vLLM APC cache
policy: exact-prefix reuse of full blocks, active-reference protection, and LRU
eviction of reusable unreferenced blocks.  It excludes vLLM's internal implementation
and additional serving optimizations. \\
lfu & Deployable & Frequency baseline with LRU tie-breaking. \\
\identifier{depth_prefer_shallow} & Deployable & Protects shallow roots. \\
\identifier{recompute_greedy} & Deployable & Protects expensive-to-recompute prefixes. \\
\identifier{cost_aware_lru} & Deployable & Blends recency and recomputation cost. \\
\identifier{prefix_fanout} & Deployable & Protects prefixes with known descendants. \\
\identifier{prefix_anchor} & Deployable & Structural anchor heuristic. \\
\identifier{tinylfu_lru} & Deployable & Selective admission with LRU-style eviction. \\
\identifier{tenant_fair_lru} & Deployable & Biases eviction toward better-served tenants. \\
\identifier{future_reuse_heuristic} & Reporting only & Uses future reuse count and next distance;
not an upper bound. \\
\identifier{oracle_future_reuse} & Reporting only & Rejects blocks with no future reuse and evicts
the legal victim whose next use is furthest away.  It is constrained by the simulator's
prefix-tree, leaf-only, and pinning rules. \\
\bottomrule
\end{longtable}

\section{State of the Art in KV-Cache Management}
\label{sec:state-of-the-art}

\subsection{There is no single ``best KV-cache algorithm''}

As of June 5, 2026, KV-cache research covers several different optimization problems
that are often reported under the same name.  Their headline speedups are not directly
comparable: papers use different models, hardware, context lengths, quality constraints,
arrival processes, baselines, and definitions of cache capacity.  A system that moves
KV tensors efficiently from SSD solves a different problem from a policy that decides
which exact-prefix block to evict from GPU memory.

The following taxonomy separates those problems before comparing their techniques:

\begin{longtable}{>{\raggedright\arraybackslash}p{0.19\textwidth}
                  >{\raggedright\arraybackslash}p{0.27\textwidth}
                  >{\raggedright\arraybackslash}p{0.25\textwidth}
                  >{\raggedright\arraybackslash}p{0.19\textwidth}}
\toprule
\textbf{Problem class} & \textbf{Representative systems} & \textbf{Primary decision}
& \textbf{Relation to this simulator} \\
\midrule
\endhead
Exact cross-request prefix reuse &
vLLM APC, SGLang RadixAttention, SAECache \cite{vllm-apc,sglang,saecache} &
Which lossless, exactly reusable prefix blocks remain in a finite accelerator cache? &
Directly in scope. \\
Workflow-aware prefix reuse &
KVFlow, INFERCEPT \cite{kvflow,infercept} &
Which paused or soon-to-run agent state should be retained, swapped, or prefetched? &
Partially modeled; workflow schedules and prefetch are absent. \\
Distributed and tiered caching &
Preble, Mooncake, MemServe, LMCache, Tutti
\cite{preble,mooncake,memserve,lmcache,tutti} &
Where should a request run, and where should its KV state live across GPU, CPU, network,
and SSD tiers? &
Out of scope for the current single-cache simulator. \\
Non-prefix and RAG reuse &
Prompt Cache, RAGCache, CacheBlend \cite{promptcache,ragcache,cacheblend} &
How can reusable modules or retrieved chunks be reused when they are not one
root-contiguous prefix? &
Out of scope; the simulator intentionally requires exact prefix contiguity. \\
Within-request KV reduction &
H2O, KIVI, SnapKV, Quest, LaProx, VeriCache
\cite{h2o,kivi,snapkv,quest,laprox,vericache} &
Which tokens, pages, layers, or tensor bits are needed while decoding one long request? &
A separate problem; it requires model-quality and attention-kernel simulation. \\
\bottomrule
\end{longtable}

For the current project, the most relevant comparison set is the first row.  The other
rows are important because they identify the next simulator dimensions, but they should
not be mixed into the current prefix-policy leaderboard.

\subsection{Production reference designs: PagedAttention, APC, and RadixAttention}

\paragraph{PagedAttention and block-based allocation.}
PagedAttention introduced operating-system-style paging for KV memory and flexible
sharing of non-contiguous physical blocks.  It substantially reduces fragmentation and
duplication, enabling larger batches and shared KV state, but it is primarily a memory
layout and allocation mechanism rather than a value-aware replacement policy
\cite{pagedattention}.  The current simulator assumes that a fixed-size block abstraction
already exists; it evaluates the policy layer built on top of that abstraction.

\paragraph{vLLM Automatic Prefix Caching.}
Current vLLM APC identifies a block by hashing its tokens together with its complete
prefix.  Only blocks with reference count zero are evictable.  The production policy
uses LRU, with deeper blocks preferred for eviction when last-access times tie
\cite{vllm-apc}.  This design preserves common ancestors and is equivalent in effect to
RadixAttention's leaf-first LRU for full-attention models.
The simulator emulates the resulting cache-policy semantics rather than vLLM's exact
implementation: full-block exact-prefix reuse, active-reference protection, and LRU
eviction of reusable unreferenced blocks.  It does not reproduce the hash-table and
free-queue data structures or additional serving optimizations such as scheduling,
allocation, continuous batching, offload, and kernels.  The comparison therefore
isolates policy logic under a common cache contract; it is not a measurement of
end-to-end vLLM throughput.

\paragraph{SGLang RadixAttention.}
RadixAttention stores reusable KV state in a radix tree, evicts the least-recently-used
leaf first, and protects nodes referenced by running requests
\cite{sglang}.  SGLang couples this cache with cache-aware scheduling and reported up to
\(6.4\times\) higher throughput than its evaluated serving baselines across structured
LLM-program workloads.  The simulator closely models RadixAttention's replacement
constraints: root-contiguous matching, leaf-only eviction, active-reference pinning,
and common-ancestor preservation.  It does not yet model the scheduler or continuous
batching effects that contribute to end-to-end SGLang performance.
The simulator treats every modeled block-tree node as a cacheable radix unit, so the
explicit \identifier{sglang_radix_attention} baseline is behaviorally equivalent to
generic leaf-LRU under this contract.  It remains visible in documentation and as an
interactive-lab option, but default comparisons exclude the duplicate row.  The
implementation mapping was checked against pinned SGLang commit
\href{https://github.com/sgl-project/sglang/tree/52f221cce088abc998fa9d3812416a45ee0e2e25/python/sglang/srt/mem_cache}{52f221cce088abc998fa9d3812416a45ee0e2e25}.
This remains a block-tree approximation: simulator capacity is charged in fixed
blocks rather than SGLang's token/page-counted radix nodes.

\paragraph{Practical conclusion.}
The production reference remains structurally constrained LRU, not a large
hand-engineered scoring function.  This is an important benchmark property: any evolved
policy should beat exact APC/Radix-style LRU by enough to justify its extra state,
scoring work, and operational complexity.

\subsection{Strongest directly relevant research policies}

\paragraph{SAECache: semantic-adaptive prefix eviction.}
The most directly relevant recent research result is SAECache, a May 2026 preprint
focused specifically on prefix-cache eviction \cite{saecache}.  Its central observation
is that blocks associated with system prompts, user queries, tool output, model
responses, and reasoning traces have very different reuse distributions.  It combines:

\begin{itemize}
  \item task-specific queues for session reuse and structural/template reuse;
  \item semantic-aware token weighting learned from eviction feedback;
  \item online adaptation of timing distributions, position decay, queue weights, and
  meta-parameters.
\end{itemize}

SAECache reports \(1.4\times\)--\(2.7\times\) TTFT improvement over production-style
baselines on its evaluated heterogeneous workloads.  It also reports that fixed
parameters can degrade by as much as \(2.7\times\) under workload mismatch.  The second
result is particularly relevant here: it supports the verifier's phase-shift, shifted
hidden families, and complexity pressure, and it argues for online adaptation rather
than a permanently tuned coefficient vector.

The current candidate cannot reproduce SAECache faithfully.  Prompt tokens are hidden,
blocks do not expose generic semantic roles, and the candidate contract has no eviction
feedback callback.  Adding those signals must preserve the trust boundary: expose
deployable generic roles such as system, user, tool, assistant, retrieved document, and
unknown, rather than exposing workload-family names or future reuse.

\paragraph{KVFlow: workflow-aware agent caching.}
KVFlow targets multi-agent workflows and represents the known execution plan as an Agent
Step Graph \cite{kvflow}.  A steps-to-execution signal guides fine-grained KV-node
eviction, while background prefetch overlaps CPU-to-GPU transfer for agents expected to
run next.  It reports up to \(1.83\times\) speedup for single large-prompt workflows and
\(2.19\times\) for concurrent workflows relative to SGLang with a hierarchical radix
cache.

This is not future leakage when the workflow graph is genuinely known to the serving
system.  It is deployable application metadata.  The current
\identifier{agent_trace_branching} workload models shared roots, branches, and retries,
but it does not expose a declared workflow graph, cache tiers, or prefetch overlap.
Consequently, the evaluator can test generic reuse learning but not KVFlow's central
algorithm.

\subsection{Systems that expand the effective cache}

Several leading systems improve KV reuse less by inventing a better single-GPU victim
score and more by changing placement, routing, or storage:

\begin{longtable}{>{\raggedright\arraybackslash}p{0.16\textwidth}
                  >{\raggedright\arraybackslash}p{0.37\textwidth}
                  >{\raggedright\arraybackslash}p{0.19\textwidth}
                  >{\raggedright\arraybackslash}p{0.17\textwidth}}
\toprule
\textbf{System} & \textbf{Core algorithmic contribution} &
\textbf{Paper-reported result} & \textbf{Missing simulator capability} \\
\midrule
\endhead
Preble &
Jointly schedules requests and prefixes across GPUs using exploitation versus
exploration, prompt-sharing features, load, and recomputation cost
\cite{preble}. &
\(1.5\times\)--\(14.5\times\) lower average latency and \(2\times\)--\(10\times\)
lower p99 latency than evaluated systems. &
Multiple workers, routing, queueing, replication, load. \\
Mooncake &
Disaggregates prefill and decode, uses GPU-cluster CPU/DRAM/SSD as a KV cache, and
schedules for effective throughput under SLOs \cite{mooncake}. &
Kimi handled 75\% more requests under reported real workloads. &
SLOs, overload rejection, prefill/decode clusters, tiers. \\
LMCache &
Provides a cache layer and control API across GPU, CPU, storage, network, engines, and
queries; pipelines and batches KV movement \cite{lmcache}. &
Up to \(15\times\) throughput improvement with vLLM in evaluated workloads. &
Bytes, transfer bandwidth, remote lookup, overlap, connectors. \\
Tutti &
Uses a GPU-centric KV object store and GPU-driven asynchronous SSD I/O with
slack-aware scheduling \cite{tutti}. &
78.3\% lower TTFT and \(2\times\) achievable request rate versus evaluated
GDS-backed SSD LMCache. &
SSD objects, I/O queues, GPU/CPU contention, slack. \\
\bottomrule
\end{longtable}

These results indicate that the globally best serving decision is often joint:
\emph{admission, eviction, tier placement, prefetch, and request routing}.  A
single-cache score can still be valuable, but the next major simulator expansion should
make transfer and placement first-class decisions.

\subsection{Beyond exact prefixes}

Exact-prefix reuse is lossless and simple, but it leaves substantial value unused in
RAG and modular prompts:

\begin{itemize}
  \item \textbf{Prompt Cache} explicitly declares reusable prompt modules and preserves
  positional correctness when those modules appear in prompts \cite{promptcache}.
  \item \textbf{RAGCache} stores retrieved knowledge states in a GPU/host knowledge tree
  and uses a RAG-aware replacement policy; it reports up to \(4\times\) lower TTFT and
  \(2.1\times\) throughput versus its evaluated vLLM--Faiss baseline
  \cite{ragcache}.
  \item \textbf{CacheBlend} reuses cached knowledge chunks even when they are not a
  prefix, selectively recomputing a small token subset to repair cross-chunk attention;
  it reports \(2.2\times\)--\(3.3\times\) lower TTFT and
  \(2.8\times\)--\(5\times\) throughput versus full KV recomputation
  \cite{cacheblend}.
\end{itemize}

Supporting these methods would require a new simulator mode.  Prefix contiguity could no
longer be the only legality rule; the model would need module identity, position
compatibility, selective recomputation cost, retrieval order, and possibly an output
quality constraint.  Those changes should not be folded silently into the existing
lossless-prefix score.

\subsection{Within-request KV reduction is a separate benchmark track}

Another major research line reduces the KV state used while decoding a single long
request.  Representative methods retain attention heavy hitters and recent tokens
(H2O), quantize keys and values asymmetrically (KIVI), select per-head important prompt
positions (SnapKV), or load query-relevant KV pages (Quest)
\cite{h2o,kivi,snapkv,quest}.  More recent preprints include LaProx, which scores token
importance using output-aware layer-wise approximation, and VeriCache, which drafts with
a compressed cache but verifies against full KV state to preserve exact outputs
\cite{laprox,vericache}.

These methods can increase effective capacity and may compose with prefix caching, but
their correctness criterion is different.  Token dropping, sparse attention, and
quantization require task-quality or exact-output verification.  They should therefore
use a separate simulator and verifier rather than competing directly against the current
lossless prefix-cache policies.

\subsection{What the current simulator covers and what to add next}

\begin{longtable}{>{\raggedright\arraybackslash}p{0.25\textwidth}
                  >{\raggedright\arraybackslash}p{0.22\textwidth}
                  >{\raggedright\arraybackslash}p{0.45\textwidth}}
\toprule
\textbf{State-of-the-art idea} & \textbf{Current coverage} &
\textbf{Required addition} \\
\midrule
\endhead
APC/Radix exact prefix matching, inactive-leaf eviction, ancestor preservation &
Strong &
Retain and continuously validate an exact production-style LRU baseline, including
deepest-prefix tie-breaking. \\
Value-aware and selective admission &
Strong &
Continue evolving compact admission and eviction scores; add value-per-byte when block
sizes become heterogeneous. \\
Semantic roles and adaptive multi-queue eviction &
Absent &
Expose generic deployable block-role metadata, add eviction-feedback hooks, and
implement a SAECache-inspired adaptive baseline. \\
Workflow proximity and prefetch &
Weak &
Add declared step graphs, pause duration, tier location, transfer time, and overlapped
prefetch events. \\
Distributed prompt-aware routing &
Absent &
Add multiple cache workers, request queues, routing, replication, per-worker load, and
network transfer. \\
GPU/CPU/SSD hierarchy &
Absent &
Replace scalar residency with tiers; model byte capacity, bandwidth, latency, pinning,
promotion, demotion, and asynchronous I/O. \\
Non-prefix module and RAG-chunk reuse &
Absent by design &
Create a separate legality and quality-preserving recomputation mode. \\
Within-request compression and sparsity &
Absent by design &
Create a separate long-context track with quality, exact-output, kernel-time, and memory
metrics. \\
\bottomrule
\end{longtable}

\paragraph{Recommended priority.}
The highest-value next comparison is a \emph{SAECache-inspired, deployable adaptive
policy} because it addresses the exact replacement problem already modeled and directly
tests whether semantic role and online adaptation beat the compact candidate.  The
second priority is a \emph{tiered workflow simulator} with declared Agent Step Graphs,
because KVFlow, LMCache, and Tutti show that prefetch and placement can dominate victim
selection.  Distributed routing and non-prefix reuse are larger, separate expansions.

\section{Evolution System}

\subsection{End-to-end workflow}

The repository wraps Levi's \code{evolve_code} API behind a stable workflow:

\begin{enumerate}
  \item \code{YamlConfigProvider} loads \filepath{configs/prefix_kv_cache.yaml}.
  \item \code{ConfigLoader} composes the problem description from the base description,
  system message, objectives, and search notes.
  \item \code{EvolutionWorkflow} writes the selected seed program to a temporary file.
  \item \code{LeviRunner} reads the seed source, creates a \code{LeviScoreFunction}, and
  calls \code{levi.evolve_code}.
  \item Levi generates and mutates candidate source under the configured evaluation
  budget.
  \item The score adapter recovers candidate source and calls
  \code{evaluate_source}; source-based evaluation ensures the complexity penalty is
  applied.
  \item The evaluator applies source-level static checks, isolates the candidate, builds
  the 90-trial train/validation selection panel plus 12 quarantined probe trials,
  returns the combined score, and preserves detailed metrics as artifacts.
  \item Levi uses \metric{combined_score} as its scalar \code{score}; successful archive
  elites also retain targeted and weakest-workload diagnostics for later mutations.
  \item After search, \code{LeviRunner} re-evaluates the best source and normalizes the
  result into \code{LeviRunResult}.
  \item The runner saves source, metrics, artifacts, metadata, summary, and a full
  baseline comparison.
\end{enumerate}

\begin{figure}[H]
\centering
\begin{tikzpicture}[
  node distance=0.55cm,
  box/.style={draw=navy, rounded corners, fill=lightblue, align=left,
              text width=12.7cm, minimum height=0.72cm},
  arrow/.style={-{Latex[length=2.3mm]}, thick, draw=darkgray}
]
  \node[box] (one) {1. Load YAML; compose prompt; select seed source.};
  \node[box, below=of one] (two) {2. Forward budget, models, pipeline, behavior,
  initialization, and punctuated-equilibrium settings to Levi.};
  \node[box, below=of two] (three) {3. Levi proposes candidate source; the adapter
  recovers source rather than only an opaque callable.};
  \node[box, below=of three] (four) {4. Isolated evaluator computes AST complexity and
  executes deterministic train/validation trials.};
  \node[box, below=of four] (five) {5. Combined score, numeric metrics, and focused
  workload diagnostics return to Levi; hidden families remain excluded.};
  \node[box, below=of five] (six) {6. Levi selects parents/inspirations, mutates again,
  and periodically applies punctuated equilibrium until the evaluation budget ends.};
  \node[box, below=of six] (seven) {7. Best source is re-evaluated and persisted with a
  full baseline report.};
  \draw[arrow] (one) -- (two);
  \draw[arrow] (two) -- (three);
  \draw[arrow] (three) -- (four);
  \draw[arrow] (four) -- (five);
  \draw[arrow] (five) -- (six);
  \draw[arrow] (six.east) -- ++(0.7,0) |- (three.east);
  \draw[arrow] (six) -- (seven);
\end{tikzpicture}
\caption{Repository-controlled evolution loop around Levi.}
\label{fig:evolution}
\end{figure}

\subsection{Current search configuration}

\begin{center}
\begin{tabularx}{\textwidth}{>{\ttfamily}p{0.30\textwidth} X}
\toprule
\normalfont\textbf{Setting} & \textbf{Current value and effect} \\
\midrule
max\_iterations & YAML default 300; CLI \code{--iterations} overrides it and is
forwarded as Levi's \code{budget_evals}. \\
mutation\_model & \code{openai/gpt-5.4-mini}, selected from the primary model. \\
paradigm\_model & \code{openai/gpt-5.5}, selected from the secondary model and
reserved for paradigm shifts. \\
sampler-model pairs & Four \code{gpt-5.4-mini} sampling configurations used by the
pipeline. \\
mutation max\_tokens & 6,000, forwarded through the normal mutation pipeline. \\
paradigm max\_tokens & 12,000 for punctuated-equilibrium reasoning calls.  A
model-specific compatibility layer replaces Levi's historical 4,096-token
paradigm sentinel; mutation calls retain their own budget. \\
n\_llm\_workers & 4. \\
n\_eval\_processes & 2, derived from \code{parallel_evaluations}. \\
eval\_timeout & 90 seconds for isolated candidate evaluation. \\
n\_parents, n\_inspirations & One parent and zero inspirations per normal mutation. \\
init & Four diverse seeds and three variants per retained seed; diverse-seed calls use
medium reasoning effort. \\
cvt & Twelve data-driven centroids, with a fallback when initialization behaviors
collapse to duplicate centroids. \\
punctuated\_equilibrium & Enabled with medium reasoning effort for
\code{gpt-5.5} paradigm-shift calls. \\
behavior & CVT descriptors include cyclomatic complexity, subscript count, validation
cache-economics metrics including avoidable eviction and short reuse after eviction,
agentic-surrogate hit and underfill, and stochastic-mix hit rate. \\
mutation feedback & Every retained parent exposes the agentic surrogate,
stochastic-serving target, and up to three additional weak-workload diagnostics to
normal mutation workers. \\
promotion gates & Agentic branching uses a fail-closed seven-metric surrogate gate
covering aggregate hit, request-tail hit, worst-quarter hit, admission waste,
underfill, short-reuse eviction misses, and normalized churn.  Cyclic working-set
pressure retains its 0.25 token-hit divergence tripwire.  Probe values remain
quarantined from selection and mutation feedback. \\
candidate guardrails & The general lane evaluates modestly over-cap candidates as
exploration-only and requests a dedicated simplification mutation; promotion remains
capped at 650 effective AST nodes.  The function-only eviction-specialist lane
evaluates ranking functions up to 1000 nodes and applies the same 650-node limit to
the final composed policy.  Rejected source patterns still fail before simulation. \\
\bottomrule
\end{tabularx}
\end{center}

\subsection{Search-operator evidence}

A completed function-only eviction-specialist run indicates that punctuated-equilibrium
proposals are high-yield structural exploration, while normal mutations provide most
cumulative improvement and the final winner.  Because this is a single-run diagnostic,
the main report treats the result as workflow evidence rather than a stable operator
benchmark.  The full sampler classification and run-level counts are retained in
\filepath{docs/research_log.tex}.

\subsection{Prompt engineering as an implementation constraint}

The system prompt documents the exact hook contract and candidate-visible schema.  It
explicitly warns candidates:

\begin{itemize}
  \item only \code{on_request_start}, \code{on_cache_hit}, and \code{on_cache_miss}
  fire as lifecycle callbacks;
  \item \code{id(block)} and guessed attributes are invalid state keys;
  \item dynamic execution, introspection, decorators, dunder access, and imports
  outside \code{math} plus the documented primitive module are rejected;
  \item future-reuse fields are unavailable to deployable candidates;
  \item recurrence, gap-moment, subtree, and regime context fields are online-only;
  \item canonical bounded multi-timescale primitives are available but optional;
  \item the compact \code{threshold_excess} positive-hinge primitive is optional;
  \item priority is useful QoS metadata but not proof of reuse;
  \item request-tail service, worst-quarter service, wasted admissions, avoidable
  evictions, churn, policy-caused underfill, and complexity affect ranking;
  \item workload-family names and request-type values must not be hard-coded;
  \item the structure probe and hidden workloads must not be inspected for candidate
  selection.
\end{itemize}

This contract is important because earlier evolved candidates repeatedly created inert
\code{on_block_*} or \code{on_request_end} machinery.  Under an unbounded complexity
penalty, unsupported hooks are not only ineffective but directly harmful.
Static and runtime contract failures now emit violation-specific repair instructions,
and mutation prompts include a preflight checklist covering syntax, entry points,
documented fields and callbacks, imports, and complexity headroom.  This keeps Python
as the structural search language while making invalid proposals cheaper to repair.
For valid retained parents, the evaluator emits at most five workload diagnostics:
the non-quarantined \identifier{agentic_tool_workflows} and
\identifier{stochastic_serving_mix} targets are reserved, and the remaining slots go
to the weakest validation workloads.  Levi samples up to three of those diagnostics
into each normal mutation prompt.  Initialization and paradigm-shift prompts still use
their separate prompt builders and do not yet receive this parent-specific feedback.

\subsection{Verifier engineering and constraint taxonomy}

The verifier is not passive measurement infrastructure; it defines the effective
optimization problem.  Building a useful verifier required substantially more than a
scalar score.  It required deterministic replay, a strict candidate-visible schema,
decision-level cache-economics audits, workload--capacity--seed decompositions,
violation-specific repair messages, weak-workload diagnostics, source-based
complexity accounting, and quarantined adjudication panels.  These diagnostics provide
the generator with fine-grained credit assignment: they identify whether a proposal
failed because of admission waste, victim choice, churn, underfill, tail service,
complexity, or an invalid implementation.  Without that decomposition, search receives
only a discontinuous ``better or worse'' signal and tends to repeat structurally
similar mistakes.

Verifier \identifier{1.0.0} is the first machine-enforced score contract.  Operative
YAML configurations declare that semantic version explicitly; evaluation results,
workload manifests, report rows, run summaries, and newly persisted evolution
snapshots repeat it beside score-bearing records.  Report builders fail closed on
missing or mixed versions before ranking or tabulation.  Historical unstamped scores
remain useful chronology, but they cannot be combined with \identifier{1.0.0}
measurements.  Configuration loading also rejects a semantic version that differs
from the verifier implemented by the checked-out source, so current code cannot
relabel its output as another verifier release.

Constraints should be classified by what a violation means, rather than making every
undesirable property an immediate rejection:

\begin{center}
\footnotesize
\begin{tabularx}{\textwidth}{@{}
  >{\raggedright\arraybackslash}p{0.19\textwidth}
  >{\raggedright\arraybackslash}p{0.31\textwidth}
  >{\raggedright\arraybackslash}X@{}}
\toprule
\textbf{Class} & \textbf{Examples} & \textbf{Search treatment} \\
\midrule
Hard feasibility and trust &
Valid Python and entry point; documented callbacks and fields; no future information
or hidden-label access; deterministic execution; legal simulator-owned transitions &
Reject before or during evaluation.  Violations make the measurement invalid, unsafe,
or non-comparable. \\
Soft performance and resource terms &
Hit rate, recompute cost, churn, wasted admission, avoidable eviction, underfill,
tail service, fairness, and effective source complexity &
Expose as metrics and penalties.  These are tradeoffs, so near-feasible candidates
retain a useful optimization gradient. \\
Staged promotion constraints &
The final 650-node deployability limit and non-regression on quarantined probe,
hidden, and tripwire panels &
Permit bounded exploration beyond the final limit, charge the excess, request an
explicit simplification mutation, then enforce the hard gate on the composed policy
before promotion. \\
\bottomrule
\end{tabularx}
\end{center}

The distinction is operational.  Making every target hard discards informative
stepping stones and turns small threshold violations into flat rejection.  Making
everything soft permits cheating, invalid measurements, and candidates that cannot be
deployed.  The practical rule is: use hard constraints for semantic validity, trust
boundaries, safety, and final production invariants; use soft constraints for
quantities whose tradeoffs should be optimized; and use staged constraints when an
over-limit proposal can contain a useful mechanism that a dedicated simplification
step may preserve.

\subsection{Experimental record}

Historical searches used several verifier revisions, so their scalar scores are not
directly comparable.  The durable workflow findings are that broad initialization can
consume the useful mutation budget, archive collapse prevents specialist retention,
and exact source plus non-seed decompositions are required to understand failed search.
The full run chronology, costs, turning points, and early coefficient-tuning record are
preserved in \filepath{docs/research_log.tex}; this report retains only current methods
and evidence-backed conclusions.

\subsection{Cost accounting and iteration velocity}

Run cost must be reported at the scope of the claim.  The \(16\)-token production
search alone cost \$4.133, but its over-cap winner was not deployable.  The required
98-evaluation simplification lineage cost another \$0.937, so the direct final-stage
cost was \$5.070 for 397 evaluations and approximately 65.0 summed run-minutes.
Claiming that the best algorithm cost less than \$5 would therefore omit a necessary
stage even before accounting for the incumbent and lessons that seeded that search.

\begin{center}
\footnotesize
\begin{tabularx}{\textwidth}{@{}
  >{\raggedright\arraybackslash}p{0.15\textwidth}
  >{\raggedright\arraybackslash}X
  rrr@{}}
\toprule
\textbf{Scope} & \textbf{Included stages} & \textbf{Evals} &
\textbf{API cost} & \textbf{Run time} \\
\midrule
Final stage &
Production search and dedicated simplification &
397 & \$5.070 & 65.0 min \\
Retained lineage &
Six discovery-verifier runs, production search, and simplification &
1,505 & \$21.981 & 221.7 min \\
Pre-promotion record &
Old-verifier iterations, structured searches, retained lineage, and both
eviction-specialist stages &
2,989 & \(\geq\)\$44.845 & \(\geq\)389.6 min \\
\bottomrule
\end{tabularx}
\end{center}

These are recorded model API charges, not total research cost.  The tally excludes
deterministic coefficient sweeps, simulator CPU time, report generation, human
analysis and composition, and experiments without reliable retained cost metadata.
The post-promotion weak-seed rediscovery run is also excluded.  Exact component values
and artifact sources are consolidated in the supplementary research log.

The stronger productivity result is \important{iteration velocity rather than
one-shot cheapness}.  Within this project, fine-grained diagnostics, persisted
candidates, explicit simplification, and automatic adjudication made new hypotheses
quick to formulate and falsify.  Failed searches did not merely consume budget: they
changed the candidate contract, archive dimensions, complexity treatment, and next
experiment.  The resulting system shortened the loop from observation to a testable
follow-up, but the cumulative cost must remain attached to the final algorithm.

\subsection{Coefficient sweep}

The reproducible tuner represents the compact policy with 15 parameters covering hit
observation weight, admission frequency/priority/structure/cost/threshold terms, and
eviction age/frequency/structure/priority terms, plus frequency and priority half-lives.
It samples a deterministic discrete grid, ranks candidates on a smaller panel, and
fully evaluates the strongest candidates.  The script defaults to 160 samples and 12
finalists; the recorded final tuning run used 180 samples and 12 finalists.  A separate
\code{--decay-ablation} mode evaluates the four combinations of enabled and disabled
frequency and priority decay on the full panel.

This hybrid process is deliberate:

\begin{itemize}
  \item LLM evolution searches policy structure and can invent state or formulas.
  \item Manual ablation removes behaviorally inert complexity.
  \item Deterministic coefficient search efficiently tunes a compact fixed structure.
\end{itemize}

\section{Compact Policy Family}

\subsection{State}

The compact policy maintains only:

\begin{itemize}
  \item a per-prefix timestamped observed frequency \(f_b\);
  \item a per-prefix timestamped maximum observed request priority \(p_b\);
  \item the current request priority.
\end{itemize}

Frequency and priority share one timestamp because both values are read and refreshed
together.  If \(\Delta t\) logical arrival steps have elapsed since the state was read,
the lazy decay operation is:

\begin{equation}
  D_h(x,\Delta t) = x\,2^{-\Delta t/h}.
\end{equation}

The selected frequency half-life is 12 steps and the selected priority half-life is
1.5 steps.  A miss adds 1.0 to the decayed frequency; a hit adds 2.5.  Priority becomes
the maximum of the decayed prior value and the current request priority.  Setting
either half-life to \code{None} disables only that decay, enabling controlled ablation.
No future information, request-type hard-coding, session map, or unsupported lifecycle
callback is used.

\subsection{Admission equation}

For block \(b\), define:

\begin{equation}
  R_b =
  0.35\log(1+\text{descendants}_b)
  -0.2\max(0,\text{depth}_b-4).
\end{equation}

The admission score is:

\begin{align}
  s_A(b) ={}&
  0.95\log(1+f_b)
  +0.2p_b
  +R_b
  +1.5\log\left(1+\frac{\text{recompute-cost}_b}{96}\right) \nonumber\\
  &-0.7
  -0.24\max(0,\text{depth}_b-2).
\end{align}

The policy therefore requires evidence before admitting most blocks, protects
structurally shared and expensive prefixes, and increasingly resists deep leaves.

\subsection{Eviction equation}

\begin{align}
  s_E(b,t) ={}&
  0.85\log(1+\max(0,t-\text{last-access}_b)) \nonumber\\
  &-1.8\log(1+f_b+\text{hit-count}_b)
  -0.2\log(1+\text{descendants}_b)
  -0.55p_b.
\end{align}

Older blocks become easier to evict, while observed reuse, global hits, structural
fanout, and priority protect a block.

\subsection{Interpretation}

The compact policy's strengths are simple and coherent:

\begin{itemize}
  \item selective admission reduces churn relative to admit-all baselines;
  \item frequency and hit weighting preserve empirically reused prefixes;
  \item recomputation cost and fanout protect valuable trunks;
  \item priority improves burst-recovery service without being the only signal;
  \item logarithms reduce sensitivity to extreme counts and ages.
\end{itemize}

The remaining stale-history term is simulator-owned \(\text{hit-count}_b\), which is
lifetime cumulative and still appears in the eviction equation.  The decayed observed
frequency partly counterbalances it, but a future ablation should test replacing the
lifetime hit count with only recent policy-owned evidence.

\subsection{Promoted pressure-aware extension}

The retained discovery incumbent in
\filepath{src/prefix_cache_evolve/problems/prefix_kv_cache/incumbents/discovery_8tok_20260608/policy.py}
preserves the compact policy's eviction equation and most admission terms.  It adds a
bounded admission-pressure state \(q_t\) derived from the online recent-admission-
pressure and miss-rate signals:

\begin{equation}
  q_t = \min\left(1.5,\;q_{t-1}2^{-1/4}
  + \text{pressure}_t\left(0.75+0.5\min(1,\text{miss-rate}_t)\right)\right).
\end{equation}

Admission subtracts a pressure penalty that increases with depth and decreases with
observed reuse and priority evidence.  The extension also uses last-access-gap evidence
and records priority only on cache hits, avoiding the assumption that a high-priority
miss predicts future reuse.  This policy was promoted only after improving selection,
aggregate probe, both recurrence-heavy probe families, and the hidden panel.
Figures~\ref{fig:incumbent-admission-first}--\ref{fig:incumbent-recurrence-frontier}
separate its admission-first organization, graduated load shedding, and narrow use of
recurrence evidence.

\begin{figure}[H]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[
  node distance=0.7cm and 0.9cm,
  evidence/.style={draw=navy, rounded corners, fill=lightblue, align=center,
                   text width=3.5cm, minimum height=1.05cm},
  regime/.style={draw=orange, rounded corners, fill=lightorange, align=center,
                 text width=3.5cm, minimum height=1.05cm},
  gate/.style={draw=navy, very thick, rounded corners, fill=graybg, align=center,
               text width=4.1cm, minimum height=1.35cm},
  admit/.style={draw=green, rounded corners, fill=lightgreen, align=center,
                text width=3.5cm, minimum height=1.05cm},
  reject/.style={draw=red, rounded corners, fill=lightorange, align=center,
                 text width=3.5cm, minimum height=1.05cm},
  note/.style={draw=darkgray, dashed, rounded corners, fill=white, align=left,
               text width=4.0cm, minimum height=0.9cm},
  arrow/.style={-{Latex[length=2.3mm]}, thick, draw=darkgray}
]
  \node[evidence] (reuse) {Observed reuse evidence\\decayed frequency and
  \code{last_access_gap}};
  \node[evidence, below=of reuse] (economics) {Prefix economics\\depth, descendants,
  recompute cost};
  \node[regime, below=of economics] (regime) {Regime context\\decayed admission pressure
  \(q_t\) and miss rate};

  \node[gate, right=1.4cm of economics] (gate) {Admission gate\\
  value \( - \) recurrence-gated depth cost\\
  \( - \) adaptive pressure costs};
  \node[admit, above right=0.5cm and 1.4cm of gate] (admit) {Positive score\\admit block};
  \node[reject, below right=0.5cm and 1.4cm of gate] (reject) {Non-positive score\\reject
  block and deeper suffix};
  \node[evidence, right=1.1cm of admit] (evict) {Stable eviction ranking\\recency, reuse,
  descendants, priority};
  \node[note, right=1.1cm of reject] (choice) {\textbf{Promoted evolved choice:}\\at
  moderate pressure, add an extra penalty only when hit-backed priority evidence is
  weak.};

  \draw[arrow] (reuse) -- (gate);
  \draw[arrow] (economics) -- (gate);
  \draw[arrow] (regime) -- (gate);
  \draw[arrow] (gate) -- node[above left]{accept} (admit);
  \draw[arrow] (gate) -- node[below left]{bypass} (reject);
  \draw[arrow] (admit) -- (evict);
  \draw[arrow, dashed] (choice) -- (reject);
\end{tikzpicture}
}
\caption{Admission-first organization of the promoted policy.  The evolved mechanisms
primarily prevent low-value blocks from entering the cache; the compact eviction
equation remains unchanged.  Priority is hit-backed evidence, not an unconditional
request-class privilege.}
\label{fig:incumbent-admission-first}
\end{figure}

\begin{figure}[H]
\centering
\begin{tikzpicture}[
  x=5.8cm,
  y=11cm,
  axis/.style={-{Latex[length=2.2mm]}, thick, draw=darkgray},
  guide/.style={densely dashed, draw=darkgray},
  low/.style={very thick, draw=orange},
  backed/.style={very thick, draw=green}
]
  \fill[lightblue, opacity=0.45] (0,0) rectangle (0.25,0.43);
  \fill[lightorange, opacity=0.55] (0.25,0) rectangle (0.8,0.43);
  \fill[red, opacity=0.08] (0.8,0) rectangle (1.5,0.43);

  \draw[axis] (0,0) -- (1.58,0)
    node[below right, text=darkgray] {decayed admission pressure \(q_t\)};
  \draw[axis] (0,0) -- (0,0.45);
  \node[rotate=90, anchor=center, text=darkgray] at (-0.16,0.225)
    {additional hinge penalty};
  \draw[guide] (0.25,0) -- (0.25,0.43);
  \draw[guide] (0.8,0) -- (0.8,0.43);

  \draw[low] (0,0) -- (0.25,0) -- (0.8,0.099) -- (1.5,0.379);
  \draw[backed] (0,0) -- (0.8,0) -- (1.5,0.154);

  \foreach \x/\label in {0/0,0.25/0.25,0.8/0.8,1.5/1.5}
    \draw[darkgray] (\x,0.006) -- (\x,-0.006) node[below] {\(\label\)};
  \foreach \y/\label in {0.1/0.1,0.2/0.2,0.3/0.3,0.4/0.4}
    \draw[darkgray] (0.012,\y) -- (-0.012,\y) node[left] {\(\label\)};

  \node[font=\scriptsize, align=center, text=navy] at (0.125,0.405)
    {ordinary\\operation};
  \node[font=\small, align=center, text=orange] at (0.525,0.405)
    {selective\\throttle};
  \node[font=\small, align=center, text=red] at (1.15,0.405)
    {persistent-pressure\\throttle};
  \node[anchor=west, text=orange] at (1.03,0.31) {no priority evidence};
  \node[anchor=west, text=green] at (1.03,0.12) {priority-backed reuse};
\end{tikzpicture}
\caption{The incumbent's graduated pressure response, isolating the two pressure-excess
hinges.  The promoted evolved term starts throttling blocks without priority evidence at
\(q_t=0.25\); the inherited all-block persistent-pressure term starts at \(q_t=0.8\).
The shared depth- and evidence-sensitive pressure penalty is omitted from this
schematic.}
\label{fig:incumbent-pressure-gates}
\end{figure}

\begin{figure}[H]
\centering
\resizebox{0.96\textwidth}{!}{%
\begin{tikzpicture}[
  node distance=0.18cm,
  admitted/.style={draw=green, rounded corners, fill=lightgreen, align=center,
                   text width=1.45cm, minimum height=0.85cm},
  rejected/.style={draw=red, rounded corners, fill=lightorange, align=center,
                   text width=1.45cm, minimum height=0.85cm},
  bypassed/.style={draw=darkgray, dashed, rounded corners, fill=graybg, align=center,
                   text width=1.45cm, minimum height=0.85cm},
  label/.style={align=right, text width=3.0cm},
  explain/.style={draw=navy, rounded corners, fill=lightblue, align=left,
                  text width=4.0cm, minimum height=1.0cm},
  arrow/.style={-{Latex[length=2.1mm]}, thick, draw=darkgray}
]
  \node[label] (coldlabel) {\textbf{First observation}\\weak recurrence evidence};
  \node[admitted, right=0.55cm of coldlabel] (c1) {depth 1\\admit};
  \node[admitted, right=of c1] (c2) {depth 2\\admit};
  \node[admitted, right=of c2] (c3) {depth 3\\admit};
  \node[rejected, right=of c3] (c4) {depth 4\\reject};
  \node[bypassed, right=of c4] (c5) {depth 5\\bypass};
  \node[bypassed, right=of c5] (c6) {depth 6\\bypass};

  \node[label, below=1.15cm of coldlabel] (repeatlabel) {\textbf{After repeated access}\\
  stronger observed reuse};
  \node[admitted, right=0.55cm of repeatlabel] (r1) {depth 1\\admit};
  \node[admitted, right=of r1] (r2) {depth 2\\admit};
  \node[admitted, right=of r2] (r3) {depth 3\\admit};
  \node[admitted, right=of r3] (r4) {depth 4\\admit};
  \node[admitted, right=of r4] (r5) {depth 5\\admit};
  \node[rejected, right=of r5] (r6) {depth 6\\reject};

  \foreach \a/\b in {c1/c2,c2/c3,c3/c4,c4/c5,c5/c6,r1/r2,r2/r3,r3/r4,r4/r5,r5/r6}
    \draw[arrow] (\a) -- (\b);

  \node[explain, right=0.9cm of c6] (coldnote) {First-time deep suffixes pay the full
  depth penalty.  One rejection blocks all deeper admission.};
  \node[explain, right=0.9cm of r6] (repeatnote) {Frequency and last-access-gap evidence
  relax depth rejection, moving the admissible frontier deeper.};
  \draw[arrow, dashed] (c6) -- (coldnote);
  \draw[arrow, dashed] (r6) -- (repeatnote);
\end{tikzpicture}
}
\caption{Recurrence-gated depth relief.  Recurrence is not a separate replacement
controller; it is a narrow exception to the default rule that deep suffixes are costly.
The exact frontier varies with block economics and pressure.}
\label{fig:incumbent-recurrence-frontier}
\end{figure}

\subsection{Decay-term ablation}

The decay half-lives were selected by screening nearby values on a small panel and then
ranking only on the full three-seed validation panel.  The following controlled
ablation uses identical policy code and excludes complexity equally from all variants:

\begin{center}
\small
\begin{tabular}{lrrrrrr}
\toprule
\textbf{Variant} & \textbf{Raw score} & \textbf{Token hit} &
\textbf{Tok. waste} & \textbf{Utility} & \textbf{Avoidable} &
\textbf{Churn/1k} \\
\midrule
No decay & 64.069 & 0.6437 & 0.5244 & 6.484 & 0.1235 & 749.8 \\
Frequency decay only & 58.160 & 0.6364 & 0.4946 & 7.108 & 0.1259 & 725.3 \\
Priority decay only & \textbf{64.839} & 0.6464 & 0.5236 & 6.499 & 0.1167 & 733.7 \\
Both decays & 64.381 & 0.6444 & 0.4913 & 7.143 & \textbf{0.1100} & \textbf{685.5} \\
\bottomrule
\end{tabular}
\end{center}

After quarantining recurrence-heavy families from selection, priority-only decay has
the highest raw validation score.  Its charged selection score is \(58.247\), compared
with \(57.788\) for both decays.  That apparent gain is a selection overfit: the
priority-only probe score falls from \(40.303\) to \(31.787\), and hidden score falls
from \(-18.616\) to \(-19.128\).  It raises probe-family token hit rates but pays for
them with worse admission/churn behavior.  The recommended compact seed therefore
retains both decays.  This inversion demonstrates why the structure probe must be
reviewed alongside the validation scalar.

\section{Results and Robustness}

\subsection{Current production-oriented result}

On the operative 16-token verifier, the promoted 572-node policy scores \(65.649\),
ahead of TinyLFU-LRU at \(63.548\).  It passes all seven checks in the agentic
surrogate gate and the cyclic working-set tripwire with a \(0.2198\) gap against a
\(0.25\) limit.  It scores \(3.064\) on the hidden panel versus \(-2.789\) for
TinyLFU-LRU and retains substantially lower churn.  The result supports two-stage
search: measured over-cap exploration can expose useful structure, but a separate
simplification phase and fail-closed adjudication are required before promotion.  The
two required runs cost \$5.070 together; the broader retained lineage
cost \$21.981.  These figures should not be collapsed into the \$4.133 production-run
cost.

\subsection{Discovery result and mechanism}

On the historical 8-token discovery verifier, the pressure-aware incumbent scores
\(77.113\), compared with \(70.362\) for TinyLFU-LRU and \(97.074\) for the
reporting-only future-reuse oracle.  Its main discovered mechanism is selective
admission under sustained pressure: low-evidence and low-priority suffixes are
throttled before they create churn, while recurrence-backed deep branches can still be
retained.  Exact ablation shows that the final priority-aware pressure term is
load-bearing; its apparent charged-probe regression is entirely complexity accounting,
while raw probe behavior improves.
The same policy behavior was originally reported as \(77.230\).  Hardening the
complexity counter raises effective complexity from 636 to 648 by charging the three
module constants, reducing only the complexity-derived score term.

\subsection{Transfer and specialist evidence}

The result does not transfer uniformly across cache geometry.  The production policy
improves the 16-token setting but over-rejects useful fine-grained blocks at 8 tokens;
at 32 tokens and above, even the future-reuse oracle is negative on the tested grid,
showing that lost prefix boundaries dominate policy choice.  Shared prefix/decode
capacity experiments likewise show that prefix eviction alone cannot repair active
decode pressure.

Focused eviction search found measurable victim-ranking improvements, but the strongest
specialist composed to a \(1{,}011\)-node policy and failed deployability.  Structured
search found richer policies that improved targeted recurrence behavior, but deletion
ablation showed broad recurrence state was harmful, priority-decay state was inert, and
complexity plus cross-panel regressions prevented promotion.  These results motivate
the retained principles below; detailed tables, per-run narratives, and negative
results are preserved in \filepath{docs/research_log.tex}, with machine-readable
evidence under \filepath{docs/results/} and \filepath{artifacts/}.

\section{Evolution-Discovered Principles}
\label{sec:evolution-principles}

The supplementary research log and machine-readable result artifacts contain more
information than the retained winner alone.  This section consolidates that evidence into a
principle ledger so future search does not repeatedly rediscover the same failures.  A
\important{retained principle} is supported by a baseline comparison, controlled
ablation, repeated search adjudication, or a combination of those sources.  The lists
below are retained within the current synthetic benchmark; statements explicitly
qualified with ``currently'' are \important{working principles} whose broader
generality remains unproven.  An \risk{open hypothesis} remains a proposed experiment
rather than a conclusion.

\subsection{Policy-design principles}

\begin{enumerate}
  \item \textbf{Selective admission is a first-order decision, not a secondary
  optimization.}  Admit-all baselines often achieve competitive hit rates, especially
  at larger capacity, but pay heavily in dead admissions and churn.  The compact
  incumbent and TinyLFU-LRU demonstrate that rejecting low-evidence suffixes can
  outperform indiscriminate retention.  Admission and eviction must therefore be
  designed together; a sophisticated victim score cannot fully repair a polluted
  cache.

  \item \textbf{Shared trunks deserve protection, while deep suffixes require stronger
  evidence.}  The compact incumbent protects fanout and recomputation cost while
  increasingly resisting deep blocks.  The structured ablation independently shows
  that removing subtree context reduces raw selection quality.  Static depth or fanout
  alone is insufficient, as the corresponding single-signal baselines trail the
  incumbent, but structural value is a useful complement to observed reuse.

  \item \textbf{Online regime context improves robustness more than headline mean
  performance.}  Removing recent admission-pressure and miss-rate conditioning raised
  the structured policy's mean workload score, but reduced raw selection, sharply
  increased churn, and damaged the aggregate probe.  Regime signals are therefore most
  valuable as control inputs that prevent unstable behavior during shifts, rather than
  as direct reuse predictors.

  \item \textbf{Explicit access-gap recurrence formulas are not automatically useful
  recurrence models.}  The tested gap-mean and gap-variance terms increased churn and
  hurt both selection and aggregate probe behavior.  Deleting them was beneficial.
  Recurrence should be represented only when it survives ablation; bounded
  multi-timescale observations are a safer search substrate than assuming that a
  hand-written gap formula is predictive.

  \item \textbf{Miss evidence is a guardrail, not a dominant value signal.}  Removing
  miss-decay state was nearly neutral on validation selection but materially weakened
  cyclic behavior.  This supports retaining a compact miss-pressure channel when it
  prevents repeated pollution or cycling failures, while rejecting large miss-state
  subsystems that do not earn their complexity.

  \item \textbf{Priority is a QoS modifier, not proof of future reuse.}  The incumbent
  uses priority only alongside frequency, structure, depth, and recomputation cost.
  The structured priority-decay channel was behaviorally inert, and the
  \identifier{priority_one_off_noise} families show why long-lived protection based
  only on priority is unsafe.  Policies should protect recurring priority traffic
  without allowing high-priority one-offs to displace reusable normal traffic.

  \item \textbf{Decay choices must be adjudicated outside the optimized split.}  In the
  compact decay ablation, priority-only decay won the validation scalar, but both-decay
  state performed better on the quarantined probe and hidden panel and reduced waste,
  avoidable eviction, and churn.  The retained lesson is not that every signal must
  decay identically; it is that stale evidence must be controlled and decay variants
  must be selected for cross-panel robustness rather than a small validation gain.

  \item \textbf{Higher hit rate can be purchased with poor cache economics.}  The
  hidden-panel compact candidate beats TinyLFU-LRU on token hit and worst-quarter hit
  yet loses badly on combined score because it admits more dead content and churns
  more.  Generated mutations repeatedly exhibit the same pattern.  A deployable policy
  must convert admitted capacity into realized saved tokens, not merely increase the
  number of matched blocks.

  \item \textbf{Richer functional form is useful only when its targeted gain survives
  operational and implementation costs.}  Generated structured mutations improved
  agent-branching behavior, and one improved raw validation plus both probe-family hit
  rates.  None survived aggregate probe, hidden, churn, waste, and complexity
  adjudication.  These are real behavioral discoveries, but they are alternate parents
  for simplification rather than deployable winners.

  \item \textbf{Compact composite scores currently generalize better than isolated
  single-signal rules.}  LRU, LFU, recomputation-only, depth-only, fanout-only, and
  fairness-biased baselines each capture a plausible principle but trail the compact
  incumbent.  The current evidence favors small formulas that combine recent reuse,
  structural value, recomputation cost, depth, and bounded priority influence.
\end{enumerate}

\subsection{Verifier and adjudication principles}

\begin{enumerate}
  \item \textbf{Hit rate is necessary but not sufficient.}  Ranking must also account
  for request tails, worst temporal windows, admission waste and utility, avoidable
  evictions, latency, churn, fairness, underfill, and implementation complexity.
  Otherwise search reliably discovers policies that buy hits through excessive
  turnover or dead retention.

  \item \textbf{Low churn is not intrinsically good.}  The compact policy can report
  very low or zero churn on difficult structure workloads by rejecting most suffixes
  and leaving capacity unused.  Policy-caused underfill is therefore required to
  distinguish valuable selectivity from broad bypass.

  \item \textbf{Realized admission outcomes should be token-weighted.}  Binary useful
  versus wasted counts are interpretable, but they treat partial and full blocks
  equally.  Token-weighted waste and saved tokens per admitted slot better represent
  whether the policy used capacity productively.

  \item \textbf{Robustness terms prevent mean-score shortcuts.}  Worst-workload,
  worst-seed, request-tail, and worst-quarter terms reveal regressions hidden by an
  improved average.  The regime ablation is the clearest example: mean performance
  improved while robust selection and operational behavior deteriorated.

  \item \textbf{Probe and hidden panels must remain quarantined from direct
  optimization.}  The decay ablation and repeated structured searches show that a
  validation winner can regress both panels.  Quarantine preserves their value as
  evidence of generalization rather than turning them into additional training splits.

  \item \textbf{Individual probe-family gains require an aggregate operational
  non-regression check.}  The strongest structured mutation improved both
  recurrence-family hit rates while its aggregate probe score deteriorated sharply.
  Per-family hits are diagnostics, not sufficient acceptance criteria.

  \item \textbf{Raw behavior and charged deployability must be reported separately.}
  Adding the complexity charge back revealed that the first expanded mutation was
  nearly tied on raw merit but decisively worse after implementation cost.  This
  decomposition distinguishes a promising functional form that needs simplification
  from a behaviorally weak idea.

  \item \textbf{Future-knowledge policies are audit instruments and ceilings, not
  deployable competitors.}  The constrained next-use oracle measures remaining
  headroom and avoidable decisions.  The weaker future-reuse heuristic also shows that
  access to future counts does not rescue a policy with poor admission and turnover
  behavior.

  \item \textbf{Multiple capacities expose adaptation failures.}  The compact
  incumbent's token-hit rate is nearly flat between 48 and 96 blocks, while several
  baselines exploit the larger cache.  Capacity sweeps prevent a policy tuned to one
  pressure regime from appearing generally optimal.
\end{enumerate}

\subsection{Evolution and search-process principles}

\begin{enumerate}
  \item \textbf{Measure discovery by learning velocity, but report cumulative cost.}
  The current policy emerged through discovery, targeted diagnostics, specialist
  experiments, over-cap exploration, and simplification.  The useful capability was
  rapid conversion of each result into a sharper experiment, not a sub-\$5 one-shot
  search.  Cost claims must include every required stage and clearly distinguish a
  final run, a retained lineage, and the broader research program.

  \item \textbf{Start from a strong incumbent and preserve budget for iteration.}  The
  first nominal 100-evaluation run spent its budget on broad initialization.  Reducing
  diverse seeds and initializing from the compact incumbent produced more useful
  iterative search.

  \item \textbf{Separate structure search, deletion ablation, and coefficient tuning.}
  LLM evolution is useful for proposing new state and formulas; controlled ablation
  identifies which terms carry value; deterministic tuning efficiently searches a
  retained compact structure.  Treating these as interchangeable wastes evaluations.

  \item \textbf{Deletion is a productive search operator.}  Manual simplification
  preserved behavior while reducing an evolved policy from 808 to 422 AST nodes.
  More importantly, the production search demonstrated the automated version: a
  656-node exploration winner triggered a dedicated simplification lineage, which
  produced deployable 539--572-node alternatives with higher charged scores.
  Structured ablation likewise improved behavior by deleting recurrence and inert
  priority state.  New state tables and branches should remain provisional until
  deletion demonstrates that they are necessary.

  \item \textbf{Complexity must be charged from exact candidate source, while safety
  and promotion caps remain separate.}  Opaque callable evaluation cannot distinguish
  a concise formula from a large defensive implementation.  Source-based, concave,
  unbounded complexity correctly breaks raw near-ties and discourages dead machinery.
  The final deployability cap should not be applied as the proposal-stage boundary:
  bounded over-cap exploration can expose useful structure, after which a dedicated
  simplification operator must recover a candidate under the strict promotion limit.

  \item \textbf{Hard and soft constraints serve different optimization roles.}
  Syntax, trust-boundary compliance, executable contracts, and final deployability are
  hard because violating them invalidates or prevents use of the result.  Cache
  economics, service quality, robustness, and exploration-stage complexity are soft
  because they are competing objectives with informative near misses.  Treating every
  target as a hard gate removes gradient; treating every invariant as a penalty admits
  invalid or undeployable winners.

  \item \textbf{Verifier engineering is part of the search algorithm.}  Deterministic
  replay and a scalar objective are necessary but insufficient.  Decision audits,
  workload-level decompositions, structured contract failures, and bounded
  parent-specific diagnostics give mutations actionable credit assignment.  The
  effort required to construct and validate those signals is a first-class part of
  the discovery cost.

  \item \textbf{The candidate contract is part of the search algorithm.}  Earlier
  candidates invented unsupported lifecycle callbacks, object-identity state, guessed
  schemas, and defensive fallback branches.  Precise prompts, stable-key guidance, and
  structured invalid results reduce this unproductive search space.

  \item \textbf{Canonical primitives require explicit discoverability and incentives.}
  Merely providing \code{MultiTimescaleDecay} did not cause viable elites to use it.
  After the prompt and seed were changed to demonstrate
  \code{observe_vector}, all 24 archived elites across three searches adopted the
  primitive.  Search representations must be easy for the generator to see and copy.

  \item \textbf{Primitive subsidies should be bounded.}  Canonical state should be
  cheaper than hand-rolled equivalents, but surrounding bespoke control flow must
  still pay its full cost.  The retained structured searches confirm that primitive
  adoption alone does not make a 1200--1400-node policy deployable.

  \item \textbf{Add primitives only after repeated successful rediscovery.}
  Separate successful searches independently found thresholded pressure-excess terms,
  justifying the small stateless \code{threshold_excess} helper.  Its one-node credit
  is intentionally smaller than the stateful decay credit, preserving pressure to
  simplify surrounding policy logic.

  \item \textbf{Repeat runs and final adjudication are required before replacing an
  incumbent.}  Three independent searches found interesting but different mutations;
  none improved charged validation while preserving both quarantined panels.  A
  single-run headline is insufficient evidence for promotion.

  \item \textbf{Persist the strongest non-seed mutation even when the seed wins.}  A
  retained incumbent can hide useful functional-form discoveries.  Saving and
  decomposing the strongest generated mutation preserves alternate parents and reveals
  whether failure came from behavior, complexity, or generalization.

  \item \textbf{Treat paradigm shifts and normal mutations as complementary operators.}
  The completed eviction-specialist diagnostic retained in
  \filepath{docs/research_log.tex} suggests that punctuated-equilibrium proposals
  are high-yield structural exploration, while normal mutations still contribute most
  cumulative best-score gain.  This supports
  alternating sparse structural exploration with dense local exploitation rather than
  replacing either operator from a single-run success-rate comparison.
\end{enumerate}

\subsection{Open hypotheses and unresolved principles}

The following items are deliberately not promoted to retained principles:

\begin{itemize}
  \item \risk{A compact form of the strongest structured mutation may retain its raw
  validation and recurrence-family gains.}  This requires deletion-oriented
  simplification and fresh probe/hidden adjudication.
  \item \risk{A compact eviction specialist may reduce avoidable and short-reuse
  eviction while preserving the full policy.}  The function-only lane isolates
  victim-ranking behavior by construction and has produced a raw specialist gain, but
  its best completed-run function composes to \(1{,}011\) nodes and no specialist has yet
  passed the complete-policy promotion gate.
  \item \risk{Lifetime \code{block.hit_count} may overprotect stale prefixes.}  The
  compact policy partly counterbalances it with decayed policy-owned frequency, but a
  controlled replacement ablation has not been run.
  \item \risk{The structure probe may work better as a lexicographic or non-regression
  constraint than as a reported-only panel.}  Directly optimizing its individual hit
  rates is already known to be unsafe; the appropriate constrained-selection mechanism
  remains untested.
  \item \risk{Capacity-aware or byte-weighted policies may change the preferred
  admission tradeoff.}  The current block-count model exposes poor capacity adaptation
  but cannot establish how byte, layer, transfer, or residency costs should alter the
  policy.
  \item \risk{Trace-calibrated and live metadata may reorder the synthetic findings.}
  The metadata-only calibration and replay path exists, but retained principles still
  require confirmation on held-out production-derived traces.
\end{itemize}

\section{Correctness Properties and Test Strategy}

\subsection{Core invariants}

Functional PyTest coverage asserts, among other properties:

\begin{itemize}
  \item hits are root-contiguous;
  \item rejected roots prevent deeper admission;
  \item missed blocks still invoke miss callbacks after deeper admission becomes
  blocked;
  \item active and interior blocks cannot be evicted;
  \item re-admitted blocks become recently used;
  \item failed lookup probes are charged;
  \item future-reuse metadata advances past the current request and handles same-step
  reuse;
  \item the oracle evicts the furthest next reuse;
  \item LRU tie-breaking and baseline distinctions behave as intended;
  \item binary and token-weighted useful/wasted admissions partition successful
  admission intervals and admitted token mass;
  \item full-block-equivalent admission utility distinguishes full and partial blocks;
  \item avoidable-eviction audits distinguish LRU from the oracle;
  \item hidden families do not affect normal combined score;
  \item structure-probe metrics and invalidity do not affect candidate selection;
  \item recurrence timestamps, bounded gap moments, subtree aggregates, and request
  regime windows are deterministic and online-computable;
  \item multi-timescale primitive state is bounded and deterministic;
  \item the smooth complexity charge is unbounded and concave;
  \item specialist evaluation freezes admission and lifecycle callbacks, ranks raw
  eviction behavior, composes winners into complete policies, and fails final
  promotion closed;
  \item form-aware complexity discounts only canonical primitive composition and leaves
  legacy mode runnable;
  \item vector observations apply distinct finite per-channel additions and reject
  mismatched widths;
  \item the selected structured seed uses bounded canonical state and the structured
  ablation harness preserves its default behavior;
  \item the quick report is labeled smoke-only;
  \item Pydantic rejects unknown workflow and evaluator settings and invalid ranges
  before execution;
  \item YAML family and score configuration matches verifier defaults;
  \item repository YAML configs explicitly declare verifier \identifier{1.0.0};
  every score tuple also carries a panel SHA-256 and an evaluation-context SHA-256
  over the verifier version, normalized evaluator configuration, and panel; report
  generation rejects missing or mixed identities within each comparison.
\end{itemize}

\subsection{Current verification status}

Under the operative verifier and workload configuration:

\begin{itemize}
  \item the full standalone repository test suite passes;
  \item \code{ruff check .} passes;
  \item TeX Live compilation succeeds;
  \item \code{git diff --check} is clean;
  \item all 29 supported workload families build;
  \item YAML/default workload and score parity is confirmed;
  \item the historical discovery baseline panel preserves credible ordering: oracle
  first, pressure-aware incumbent second, TinyLFU-LRU third, the reporting heuristic
  fourth, and prefix-anchor and vLLM APC following;
  \item on the operative 16-token default, the promoted production incumbent leads
  TinyLFU-LRU by \(2.101\) charged points (\(65.649\) versus \(63.548\)), while
  retaining higher validation token hit and much lower churn.
\end{itemize}

\section{Limitations and Recommended Next Steps}

\subsection{Simulator limitations}

\begin{itemize}
  \item \textbf{Block capacity rather than byte capacity.} All blocks consume one unit
  despite possible variation in layer count, dtype, model width, or partial-block
  memory.
  \item \textbf{Proxy latency.} The deterministic formula is useful for ranking but does
  not include kernel scheduling, bandwidth, allocator overhead, or device contention.
  \item \textbf{Synthetic traffic.} Workloads encode plausible failure modes but are not
  calibrated from production trace distributions.
  \item \textbf{Single-cache topology.} There is no routing, replication, cache
  partitioning, host offload, or distributed ownership.
  \item \textbf{Shared-KV mode is a demand stress test, not a scheduler.} It exposes
  infeasible decode demand and prefix displacement, but does not model request
  rejection, preemption, swapping, batch formation, or decode slowdown after an
  allocation failure.
  \item \textbf{Exact prefix reuse only.} There is no semantic reuse, chunk reuse outside
  prefix position, or tokenizer/model-version mismatch.
  \item \textbf{Token rather than byte utility.} Admission utility distinguishes token
  count and repeated hits but does not yet model model-dependent bytes, residency time,
  or transfer cost.
  \item \textbf{Audit horizon.} Avoidable eviction uses full future knowledge for
  verification, which is appropriate for regret measurement but should not be confused
  with an implementable signal.
\end{itemize}

\subsection{Verifier limitations}

\begin{itemize}
  \item Score weights encode design preferences and can still steer search toward the
  cheapest gradient.  They should be sensitivity-tested.
  \item Fairness cost currently uses only the \code{multi_tenant_skew} workload even
  though tenant metrics are reported elsewhere.
  \item Hidden workloads are quarantined from selection, but they remain synthetic and
  visible in repository source to human developers.
  \item The minimum workload term can be dominated by one structurally difficult family;
  score breakdowns must be reviewed alongside the scalar.
  \item Current recovery metrics recognize request types containing the word
  \code{recovery}; this is acceptable as verifier instrumentation but is not a general
  production phase detector.
\end{itemize}

\subsection{Evolution limitations}

\begin{itemize}
  \item The LLM search can invent unsupported or dead state despite prompt guidance.
  Complexity discourages this but does not prevent it.
  \item Fine-grained evaluator diagnostics currently guide normal mutations only.
  Initialization and paradigm shifts do not receive parent-specific diagnostics, and
  the workflow has no bounded global memory of rejected ideas or recurring behavioral
  failures.  The eviction-specialist lane narrows the search problem but does not
  remove this limitation.
  \item A full candidate evaluation is relatively expensive, limiting broad structural
  search.
  \item The coefficient tuner optimizes a fixed compact structure and cannot discover
  qualitatively new state representations.
  \item Three additional 98-evaluation searches from the ablated structured seed did
  not beat the compact seed.  Canonical primitive adoption improved, but
  generated programs still accumulated substantial bespoke control flow.
  \item A later 298-evaluation compact-family search promoted the pressure-aware
  incumbent, which improved selection, probe, and hidden adjudication.
\end{itemize}

\subsection{Recommended next experiments}

\begin{enumerate}
  \item Compact or structurally ablate the strongest mutation from run
  \identifier{20260606T125829Z}.  It is the first generated form to improve raw
  validation and both recurrence-family hit rates, but its \(1322\)-node effective
  implementation and aggregate-probe regression prevent acceptance.
  \item Test an archive dimension or search-time non-regression constraint for
  operational probe score.  Final specialist promotion already enforces aggregate and
  per-family probe non-regression; the unresolved question is whether earlier
  constrained retention improves search without leaking the probe into the optimized
  scalar.
  \item Decompose and simplify the completed function-only winner, then replicate the
  eviction-specialist search to test whether raw gains can survive the 650-node
  composed-policy promotion gate.
  \item Replace block-count capacity with byte- or layer-weighted capacity and add
  byte-weighted admission utility.
  \item Extend the shared prefix/decode capacity panel with scheduler actions and an
  explicit decode-failure or SLO objective.  The current reasoning panel's \(77.5\%\)
  incumbent allocation-failure rate shows that eviction-only optimization is the wrong
  control surface under sustained reasoning bursts.
  \item Calibrate workload mixture, prefix depths, arrival bursts, output lengths, and
  tenant phase durations from anonymized production traces, then replay held-out
  trace-derived metadata.
  \item Add routing, tiering, and host-offload simulators so placement policies can be
  compared with single-cache victim scoring.
  \item Add a bounded global failure store and feed recurring repair and behavioral
  failures to initialization and paradigm-shift prompts, which currently lack the
  parent-specific diagnostics available to normal mutations.
  \item Revisit lifetime \code{block.hit_count}; it remains stale permanent state in the
  compact eviction score even after policy-owned frequency decay.
\end{enumerate}

\section{Reproducibility}

\subsection{Run the full baseline report}

\begin{lstlisting}[style=shell]
.venv/bin/python -m prefix_cache_evolve.problems.prefix_kv_cache.runner \
  --baseline-report \
  --candidate-program \
  src/prefix_cache_evolve/problems/prefix_kv_cache/incumbents/production_16tok_20260609/policy.py
\end{lstlisting}

\subsection{Reproduce the historical discovery result and workload panel}

\begin{lstlisting}[style=shell]
.venv/bin/prefix-cache-evolve \
  --workload-manifest \
  --config configs/prefix_kv_cache_discovery.yaml \
  --workload-manifest-output /tmp/discovery_workload_manifest.json

.venv/bin/prefix-cache-evolve \
  --baseline-report \
  --candidate-program \
  src/prefix_cache_evolve/problems/prefix_kv_cache/incumbents/discovery_8tok_20260608/policy.py \
  --config configs/prefix_kv_cache_discovery.yaml
\end{lstlisting}

\subsection{Run the quarantined reports}

\begin{lstlisting}[style=shell]
.venv/bin/python -m prefix_cache_evolve.problems.prefix_kv_cache.runner \
  --probe-report \
  --candidate-program \
  src/prefix_cache_evolve/problems/prefix_kv_cache/incumbents/production_16tok_20260609/policy.py

.venv/bin/python -m prefix_cache_evolve.problems.prefix_kv_cache.runner \
  --hidden-report \
  --candidate-program \
  src/prefix_cache_evolve/problems/prefix_kv_cache/incumbents/production_16tok_20260609/policy.py
\end{lstlisting}

\subsection{Run a new evolution from the production incumbent}

\begin{lstlisting}[style=shell]
.venv/bin/python -m prefix_cache_evolve.problems.prefix_kv_cache.runner \
  --iterations 300 \
  --config configs/prefix_kv_cache.yaml \
  --seed-program \
  src/prefix_cache_evolve/problems/prefix_kv_cache/incumbents/production_16tok_20260609/policy.py
\end{lstlisting}

\subsection{Run eviction-specialist exploration}

\begin{lstlisting}[style=shell]
.venv/bin/python -m prefix_cache_evolve.problems.prefix_kv_cache.runner \
  --iterations 300 \
  --config configs/prefix_kv_cache_eviction_specialist.yaml \
  --artifact-output artifacts/prefix_kv_cache_eviction_specialist_runs
\end{lstlisting}

\subsection{Run the structured ablation and search parent}

\begin{lstlisting}[style=shell]
.venv/bin/prefix-cache-tools ablate structured

.venv/bin/python -m prefix_cache_evolve.problems.prefix_kv_cache.runner \
  --iterations 100 \
  --config configs/prefix_kv_cache.yaml \
  --seed-program \
  src/prefix_cache_evolve/problems/prefix_kv_cache/seeds/structured_seed.py
\end{lstlisting}

\subsection{Run the deterministic compact-policy tuner}

\begin{lstlisting}[style=shell]
.venv/bin/prefix-cache-tools tune compact \
  --samples 180 \
  --full-top 12 \
  --seed 20260605
\end{lstlisting}

\subsection{Run tests and lint}

\begin{lstlisting}[style=shell]
.venv/bin/pytest -q
.venv/bin/ruff check .
\end{lstlisting}

\subsection{Run eviction decision and distillation analysis}

\begin{lstlisting}[style=shell]
.venv/bin/prefix-cache-tools analyze eviction
\end{lstlisting}

\subsection{Run shared reasoning decode-KV robustness analysis}

\begin{lstlisting}[style=shell]
.venv/bin/prefix-cache-tools analyze reasoning-kv
\end{lstlisting}

\subsection{Artifact layout}

Each saved run directory contains core artifacts and may contain the applicable
run-type-specific artifacts:

\begin{itemize}
  \item \filepath{best_program.py}: best source returned by Levi;
  \item \filepath{metrics.json}: flattened evaluator metrics;
  \item \filepath{artifacts.json}: split, workload, capacity, metadata, and score
  breakdown artifacts;
  \item \filepath{metadata.json}: Levi cost and runtime metadata;
  \item \filepath{run_summary.json}: evaluations, cost, runtime, archive size, and seed;
  \item \filepath{config_snapshot.yaml}: operative search configuration, when the run
  was launched from a config file;
  \item \filepath{workload_manifest.json}: deterministic stream summaries and panel
  fingerprint for the operative report configuration;
  \item \filepath{surrogate_probe_tripwires.json} and
  \filepath{surrogate_probe_tripwires.md}: fail-closed gate suite containing the
  multi-metric agentic comparison and cyclic working-set token-hit tripwire;
  \item \filepath{agentic_surrogate_probe_gate.json} and
  \filepath{agentic_surrogate_probe_gate.md}: detailed per-metric agentic gate result;
  \item \filepath{promotion_adjudication.json} and
  \filepath{promotion_adjudication.md}: fail-closed composed complete-policy
  adjudication for specialist runs;
  \item \filepath{promotion_candidate.py}: complete composed policy evaluated by
  function-only specialist promotion;
  \item \filepath{seed_program.py}: exact source used as the search parent when a
  generated non-seed mutation is retained;
  \item \filepath{best_generated_mutation.py} and
  \filepath{best_generated_mutation_snapshot.json}: strongest archived generated
  non-seed, even when the seed remains the returned winner;
  \item \filepath{best_generated_mutation_decomposition.json} and
  \filepath{best_generated_mutation_decomposition.md}: selection, probe, hidden, and
  complexity comparison between the seed and strongest generated mutation, when one is
  available;
  \item \filepath{paradigm_candidates/}: copied paradigm-shift candidates, when Levi
  retained them;
  \item \filepath{baseline_comparison.md}: full candidate-versus-baseline report, or
  \filepath{baseline_comparison_error.json} if report generation failed.
\end{itemize}

\section{Conclusion}

The repository combines a deterministic prefix-tree simulator, a robust verifier, and
an LLM-guided source-evolution workflow.  Its central systems result is that admission,
eviction, and implementation complexity must be evaluated together: higher hit rate
alone frequently purchases excessive churn, dead admissions, or poor transfer.

The strongest evolved policies use compact pressure-aware admission with bounded
structural and recurrence evidence.  The successful search process also depends on
fine-grained evaluator feedback, specialist archive dimensions, explicit
simplification, exact-source complexity, and quarantined probe and hidden panels.
Its main research-productivity lesson is that instrumented source evolution can make
the propose--test--diagnose--simplify loop substantially faster inside a project.
That speed produced lessons and follow-up experiments quickly; it did not make the
cumulative discovery program a sub-\$5 exercise.
Current evidence supports these design principles, but production-trace transfer,
shared decode-capacity control, and geometry-robust promotion remain open.

Detailed chronology, experiment tables, and rejected solutions are intentionally
consolidated in \filepath{docs/research_log.tex}; generated measurements remain under
\filepath{docs/results/} and \filepath{artifacts/}.  This keeps the main report focused
on the current method, claims, limitations, and reproducibility contract.

\appendix

\section{Supplementary Experimental Record}

The detailed engineering journey, run diary, historical verifier results, structured
search detours, and rejected evolutionary solutions are retained in the separately
compilable \filepath{docs/research_log.tex}.  Machine-readable manifests and score
matrices remain under \filepath{docs/results/}, and exact run artifacts remain under
\filepath{artifacts/}.
Separating that evidence record keeps this report focused while preserving the
provenance needed to audit or reinterpret its conclusions.

\section{A Mathematical Model of Regime-Conditional Admission--Eviction Control}
\label{app:admission-mathematical-model}

\subsection{Status}

The experiments suggest the following hypothesis:

\begin{quote}
\emph{For the current \identifier{pressure\_aware\_incumbent}, admission mistakes are
larger in aggregate under the future-occurrence surrogate audit, but eviction mistakes
are the binding error mode in identifiable workload regimes.  The useful mathematical
object is not an admission-only theory; it is a regime-conditional model of admission
classification, eviction ranking, and their interaction under prefix-tree legality and
complexity constraints.}
\end{quote}

This statement separates four facts that should not be collapsed:

\begin{enumerate}
  \item \textbf{Local surrogate regret.}  The audit assigns future-occurrence value to
  blocks and compares the incumbent decision with a same-state local alternative.
  Under this surrogate, aggregate admission-side regret is much larger than aggregate
  eviction-side regret.
  \item \textbf{Groupwise heterogeneity.}  The universal dominance claim is false:
  eviction-dominant counterexamples occur in heavy-tailed, shared-root, RAG-template,
  priority-noise, and multi-tenant regimes.
  \item \textbf{Eviction opportunity.}  In the incumbent replay, \(77.5\%\) of
  \(5{,}402\) eviction decisions have multiple legal victims, with 10.64 legal victims
  on average.  A full specialist changes the selected victim on \(21.5\%\) of eviction
  decisions.
  \item \textbf{Causal component value.}  In the \(2\times2\) incumbent/oracle crossed
  replay, admission has \(3.29\times\) the aggregate main effect of eviction.
  Oracle admission removes \(81.1\%\) of eviction value in the groups that the local
  audit labeled eviction-dominant, although \(37/162\) groups retain positive residual
  eviction value.
  \item \textbf{Deployability and complexity.}  The best completed function-only
  specialist improves raw specialist score from \(70.989\) to \(72.091\), but composes
  to a \(1{,}011\)-node complete policy.  The model must therefore distinguish
  behavioral headroom from complexity-constrained promotability.
\end{enumerate}

The main modeling correction is to replace a uniform admission-dominance claim with a
conditional one.  Admission dominates when threshold mistakes are frequent or when
low-value admitted blocks pollute the future state.  Eviction dominates when the legal
victim set has high value spread and the ranker chooses an expensive leaf.

\begin{figure}[H]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[
  state/.style={draw=navy, very thick, rounded corners, fill=lightblue,
                align=center, text width=3.55cm, minimum height=1.35cm},
  quantity/.style={draw=orange, rounded corners, fill=lightorange,
                   align=center, text width=4.05cm, minimum height=1.45cm},
  control/.style={draw=green, rounded corners, fill=lightgreen,
                  align=center, text width=4.15cm, minimum height=1.45cm},
  decision/.style={draw=orange, diamond, fill=graybg, align=center,
                   aspect=2.15, text width=2.5cm},
  outcome/.style={draw=navy, rounded corners, fill=graybg,
                  align=center, text width=3.7cm, minimum height=1.25cm},
  audit/.style={draw=darkgray, dashed, rounded corners, fill=white,
                align=center, text width=4.0cm, minimum height=1.15cm},
  arrow/.style={-{Latex[length=2.4mm]}, thick, draw=darkgray},
  audit-arrow/.style={-{Latex[length=2.2mm]}, thick, dashed, draw=darkgray}
]
  \node[state] (state) at (0,0) {\textbf{Feasible cache state}\\
    \(X_t=(S_t,q_t,D_t,Z_t)\)\\
    ancestor closed; capacity bounded\\
    legal victims \(\mathcal L(X_t)\) are inactive leaves};

  \node[quantity] (admission) at (4.8,2.15) {
    \textbf{Admission classification}\\
    incoming value \(F_t(b)\) versus displacement \(\mu_t\)\\
    margin \(\gamma^A_t(b)=|F_t(b)-\mu_t|\)\\
    large wrong-side margin \(\Rightarrow\) large \(R^A_t(b)\)};

  \node[quantity] (eviction) at (4.8,-2.15) {
    \textbf{Eviction opportunity}\\
    victim count \(n^E_t=|\mathcal L(X_t)|\)\\
    value spread \(\delta^E_t=\max_{u,v}|F_t(u)-F_t(v)|\)\\
    chosen-victim excess \(\Rightarrow\) \(R^E_t\)};

  \node[control] (admit-control) at (10.0,2.15) {
    \textbf{Selective admission control}\\
    estimate \(\widehat G_t(b)-\lambda_t w_b\)\\
    admit iff score \(>0\); rejection blocks deeper suffix\\
    pressure raises the effective threshold};

  \node[decision] (gate) at (10.0,-2.15) {
    \(n^E_t\ge2\) and\\
    estimated victim\\
    spread is large?};

  \node[control] (evict-control) at (15.5,-2.15) {
    \textbf{Regime-gated eviction control}\\
    \(h_t=0\): compact frequency-aware ranker\\
    \(h_t=1\): richer victim-value ranker\\
    hybrid score in Equation~\eqref{eq:regime-gated-eviction}};

  \node[outcome] (transition) at (20.3,1.1) {
    \textbf{Legal simulator transition}\\
    \(X_{t+1}=\Phi(X_t,r_t,\pi)\)\\
    pinning, leaf eviction, rollback, capacity};

  \node[audit] (evaluation) at (13.5,-5.45) {
    \textbf{Two evidence levels}\\
    local audit: \(A_g,E_g,\bar A_g,\bar E_g\)\\
    crossed replay: realized hit, churn, tails, and score};

  \node[audit] (deployability) at (19.5,-5.45) {
    \textbf{Staged constraints}\\
    soft charge \(+\) bounded over-cap search\\
    simplify, then hard promotion at \(\operatorname{cx}(\pi)\le650\)};

  \draw[arrow] (state) -- (admission);
  \draw[arrow] (state) -- (eviction);
  \draw[arrow] (admission) -- (admit-control);
  \draw[arrow] (eviction) -- (gate);
  \draw[arrow] (gate) -- (evict-control);
  \draw[arrow] (admit-control) -- (transition);
  \draw[arrow] (evict-control.east) -- (transition.south west);
  \draw[audit-arrow] (transition.south) |- (evaluation.east);
  \draw[audit-arrow] (evaluation) -- (deployability);
\end{tikzpicture}
}
\caption{Regime-conditional admission--eviction model.  The same feasible prefix-tree
state induces two distinct error opportunities: threshold classification for an
incoming block and ranking among legal victims.  Selective admission is always active;
additional eviction sophistication is gated by victim multiplicity and estimated value
spread.  Local future-occurrence regret diagnoses the error class, crossed replay
measures realized trajectory effects, while staged complexity treatment preserves
over-cap discoveries without relaxing the final deployability requirement.}
\label{fig:regime-conditional-mathematical-model}
\end{figure}

\subsection{Trace, prefix tree, and feasible states}

Fix a finite request trace \(\omega=(r_1,\ldots,r_T)\).  Request \(r_t\) has tenant
identifier \(\tau_t\), prompt token sequence
\(x_t=(x_{t,1},\ldots,x_{t,L_t})\), priority \(\rho_t\), arrival metadata, and true
output length.  For block size \(B\), define the tenant-qualified cumulative-prefix
chain
\begin{equation}
  \mathbf b_t=(b_{t,1},\ldots,b_{t,n_t}),
  \qquad
  n_t=\left\lceil \frac{L_t}{B}\right\rceil,
\end{equation}
where \(b_{t,d}\) is the block representing the first \(\min(dB,L_t)\) prompt tokens.
Let \(\operatorname{pa}(b)\) be the parent block, \(\operatorname{Anc}(b)\) the strict
ancestors, \(d(b)\) the depth, \(s_b\) the token count, and \(w_b>0\) the capacity
weight.  The current simulator uses \(w_b=1\) for prefix blocks, while token-weighted
metrics use \(s_b\); a byte-accurate extension can replace \(w_b\) by KV bytes.

At logical time \(t\), before processing request \(r_t\), write the simulator state as
\begin{equation}
  X_t=(S_t,q_t,D_t,Z_t),
\end{equation}
where \(S_t\) is the resident prefix-block set, \(q_t(b)\in\mathbb N\) is the active
pin count for block \(b\), \(D_t\ge 0\) is active decode-KV occupancy in shared-capacity
experiments, and \(Z_t\) is policy-owned online state.  A physical state is feasible
when
\begin{align}
  b\in S_t &\Longrightarrow \operatorname{Anc}(b)\subseteq S_t,
  \label{eq:prefix-closure-regime}\\
  q_t(b)>0 &\Longrightarrow b\in S_t,
  \label{eq:pins-resident-regime}\\
  \sum_{b\in S_t} w_b + D_t &\le C.
  \label{eq:capacity-regime}
\end{align}
Thus the prefix cache is an ancestor-closed, dynamically capacitated tree subset.

The legal victim set is
\begin{equation}
  \mathcal L(X_t)=
  \left\{b\in S_t:
  q_t(b)=0\;\land\;
  \{c\in S_t:\operatorname{pa}(c)=b\}=\varnothing
  \right\}.
  \label{eq:legal-victims-regime}
\end{equation}
Only inactive resident leaves are evictable.  Interior nodes preserve reachability of
resident descendants, and pinned nodes are referenced by active requests.

\subsection{Online policies, frozen components, and complexity-constrained classes}

Let \(\mathcal F_t\) be the candidate-visible history through time \(t\): scrubbed
request metadata, candidate-visible block metadata, lifecycle callbacks, previous
candidate decisions, and candidate-owned state.  It excludes future requests, raw
prompt tokens, hidden workload labels, true future reuse, and simulator-only output
lengths.

A deployable policy is an \(\mathcal F_t\)-adapted tuple
\begin{equation}
  \pi=(\pi_{\mathrm{start}},\pi_{\mathrm{hit}},\pi_{\mathrm{miss}},
       \alpha,\varepsilon),
\end{equation}
where \(\alpha_t(b)\) is the admission score and \(\varepsilon_t(b)\) is the eviction
score.  The simulator converts scores into actions by
\begin{align}
  \text{admit } b &\Longleftrightarrow \alpha_t(b)>0,\\
  e_t &\in \arg\max_{b\in\mathcal L(X_t)}\varepsilon_t(b),
\end{align}
with deterministic tie-breaking and legality enforcement owned by the simulator.

The eviction-specialist lane is naturally represented as a restricted component class.
Let
\begin{equation}
  \pi_0=(\pi_{0,\mathrm{start}},\pi_{0,\mathrm{hit}},\pi_{0,\mathrm{miss}},
       \alpha_0,\varepsilon_0)
\end{equation}
be the pressure-aware incumbent.  The specialist search freezes
\((\pi_{0,\mathrm{start}},\pi_{0,\mathrm{hit}},\pi_{0,\mathrm{miss}},\alpha_0)\) and
searches only an eviction scorer \(\varepsilon_\phi\), producing composed policies
\begin{equation}
  \pi(\phi)=
  (\pi_{0,\mathrm{start}},\pi_{0,\mathrm{hit}},\pi_{0,\mathrm{miss}},
   \alpha_0,\varepsilon_\phi).
\end{equation}

Let \(\operatorname{cx}(\pi)\) be the effective AST complexity after canonical-primitive
subsidies.  The deployable class is not merely behavioral; it is complexity constrained:
\begin{equation}
  \Pi_\kappa=\{\pi:\pi\text{ is deployable and }\operatorname{cx}(\pi)\le \kappa\},
  \qquad \kappa=650
\end{equation}
for the hard promotion gate.  The verifier also uses a smooth charged objective with a
complexity penalty.  Consequently, an eviction scorer can reveal real behavioral
headroom while still failing promotion because its composed policy lies outside
\(\Pi_\kappa\) or loses on hidden generalization.

\subsection{Simulator transition and root-contiguous admission}

For resident set \(S_t\), the root-contiguous hit length of request \(r_t\) is
\begin{equation}
  m_t(S_t)=\max\{k\le n_t:b_{t,1},\ldots,b_{t,k}\in S_t\}.
\end{equation}
Blocks \(b_{t,1},\ldots,b_{t,m_t}\) are prompt-prefix hits and are pinned for the
request.  Blocks \(b_{t,m_t+1},\ldots,b_{t,n_t}\) are recomputed misses.  Admission is
attempted sequentially along this missed suffix.  A block can be admitted only if its
parent remains resident; a rejection prevents deeper admission from being scored;
a tentative admission is pinned before capacity enforcement; capacity enforcement
repeatedly removes the legal victim with maximum \(\varepsilon_t\); and if no legal
victim exists while the state remains over capacity, the tentative block is rolled back
and the remaining suffix is forced to bypass.

With fixed trace, tie-breaking, and policy source, this defines a deterministic
transition
\begin{equation}
  X_{t+1}=\Phi(X_t,r_t,\pi).
\end{equation}
All reported quantities--token hit, block hit, recompute cost, latency proxy,
underfill, churn, admission waste, avoidable eviction, and fairness--are functionals of
the realized trajectory
\((X_1,r_1,X_2,r_2,\ldots,X_{T+1})\).

\subsection{Objective, offline comparator, and local value relaxation}

For workload distribution \(\mathcal P\) and deployable class \(\Pi_\kappa\), the ideal
policy problem is
\begin{equation}
  \pi^*\in\arg\max_{\pi\in\Pi_\kappa}
  \mathbb E_{\omega\sim\mathcal P}[J_\theta(\pi;\omega)],
  \label{eq:ideal-regime-objective}
\end{equation}
where \(J_\theta\) is the verifier-induced serving-quality functional.  The implemented
verifier replaces \(\mathcal P\) with a deterministic panel of workload families,
capacities, and seeds:
\begin{equation}
  \widehat J_\theta(\pi)=
  \Psi_\theta\left(\{M(\pi;\omega_{f,c,s})
  : f\in\mathcal F_{\mathrm{val}}, c\in\mathcal C, s\in\mathcal S\}\right)
  -0.065\operatorname{cx}(\pi)^{0.75}.
  \label{eq:empirical-regime-objective}
\end{equation}
Here \(M\) is the metric vector from the deterministic simulator and \(\Psi_\theta\) is
the aggregation formula that blends mean performance, weak-workload performance,
worst-seed behavior, cache-economics charges, and service-quality penalties.

More generally, let \(h_i(\pi)\le0\) denote hard feasibility constraints and let
\(\ell_j(\pi)\ge0\) denote soft violation magnitudes.  Hard constraints include the
candidate contract, information boundary, execution limits, and simulator legality.
Soft terms include cache-economics losses and robustness deficits; source complexity
is charged separately.  Let \(U_\theta(\pi)\) denote the uncharged serving utility
aggregated from the validation panel.  The search problem is therefore
\begin{align}
  \Pi_{\mathrm{hard}}
    &=\{\pi:h_i(\pi)\le0,\ i=1,\ldots,m\},\\
  \pi_{\mathrm{search}}
    &\in\arg\max_{\substack{\pi\in\Pi_{\mathrm{hard}}\\
                            \operatorname{cx}(\pi)\le\kappa_{\mathrm{explore}}}}
      \left[
        U_\theta(\pi)
        -\sum_{j=1}^{n}\lambda_j\ell_j(\pi)
        -0.065\operatorname{cx}(\pi)^{0.75}
      \right],
  \label{eq:hard-soft-search}
\end{align}
where \(\kappa_{\mathrm{explore}}>\kappa\) is a bounded outer guard rather than the
production limit.  The outer guard prevents runaway proposals; it does not force every
useful intermediate representation to satisfy the final cap.

For an over-cap candidate, a simplification operator \(\mathcal S\) searches for a
behaviorally equivalent or improved descendant.  Let
\(\mathcal N_{\mathrm{delete}}(\pi)\) denote deletion-oriented and algebraic
simplifications of \(\pi\):
\begin{equation}
  \mathcal S(\pi)
  \in\arg\max_{\pi'\in\mathcal N_{\mathrm{delete}}(\pi)}
  \left[
    U_\theta(\pi')-\sum_j\lambda_j\ell_j(\pi')
    -0.065\operatorname{cx}(\pi')^{0.75}
  \right]
  \quad\text{subject to}\quad
  \pi'\in\Pi_{\mathrm{hard}},
  \operatorname{cx}(\pi')\le\kappa.
  \label{eq:simplification-operator}
\end{equation}
Promotion then applies hard production and adjudication gates:
\begin{equation}
  \Pi_{\mathrm{promote}}=
  \left\{\pi\in\Pi_{\mathrm{hard}}:
    \operatorname{cx}(\pi)\le650,
    Q_k(\pi)\ge Q_k(\pi_0)-\epsilon_k\ \forall k
  \right\},
  \label{eq:promotion-class}
\end{equation}
where \(Q_k\) are the quarantined probe, hidden, and tripwire criteria and \(\pi_0\) is
the incumbent.  These \(Q_k\) values adjudicate promotion but are not returned as
mutation targets.  This separation preserves their evidentiary role while allowing
soft validation feedback to guide search.

A clairvoyant but still legal offline comparator can be defined by dynamic programming
on an additive surrogate reward \(g_\theta\):
\begin{align}
  V_{T+1}(X)&=0,\\
  V_t(X)&=\max_{u\in\mathcal U(X,r_t)}
  \{g_\theta(X,u,r_t)+V_{t+1}(\Phi(X,r_t,u))\},
  \label{eq:offline-regime-dp}
\end{align}
where \(\mathcal U(X,r_t)\) contains only action sequences obeying prefix-contiguous
admission, inactive-leaf eviction, pinning, rollback, and capacity.  This DP is
infeasible at report scale, but it defines the rigorous object approximated by
future-aware oracles and rollout audits.

A practical online policy instead estimates marginal block value.  Let \(G_t(b)\) be
the conditional gross benefit of retaining block \(b\) under future requests, and let
\(\lambda_t\) be a capacity shadow price.  The local relaxation is
\begin{equation}
  \widehat\Delta_t(b)=\widehat G_t(b)-\lambda_t w_b.
  \label{eq:local-shadow-regime}
\end{equation}
The relaxed first-order rules are
\begin{align}
  \text{admit } b
  &\Longleftrightarrow
  \widehat G_t(b)>\lambda_t w_b
  \quad\text{and}\quad \operatorname{pa}(b)\in S_t,
  \label{eq:relaxed-admission-regime}\\
  \text{evict } e_t
  &\in\arg\min_{b\in\mathcal L(X_t)}
  \frac{\widehat G_t(b)}{w_b}.
  \label{eq:relaxed-eviction-regime}
\end{align}
Frequency, recent recurrence, subtree support, priority-backed reuse, and recomputation
cost estimate \(\widehat G_t\).  Admission pressure, miss pressure, occupancy, and
shared decode demand estimate \(\lambda_t\).  The current pressure-aware policy mainly
changes the admission threshold, while the specialist lane asks whether a richer
estimator of \(\widehat G_t\) improves the eviction argmin.

\subsection{Future-occurrence audit as a local same-state decomposition}

The decision audit uses a stable future-value surrogate rather than a full causal
counterfactual.  Let
\begin{equation}
  F_t(b)=s_b\,N_t^{\mathrm{future}}(b)
\end{equation}
be block token count times the number of remaining request occurrences after the
current request.  For a reachable incoming block \(b\), define the local displacement
threshold
\begin{equation}
  \mu_t=
  \begin{cases}
    0, & \text{if enough free prefix capacity exists},\\
    \min_{v\in\mathcal L(X_t)} F_t(v), & \text{otherwise},
  \end{cases}
  \label{eq:local-displacement-threshold}
\end{equation}
with the convention that admission is infeasible if capacity is required and
\(\mathcal L(X_t)=\varnothing\).

The one-step admission-side surrogate regret for scored reachable block \(b\) is
\begin{equation}
  R^A_t(b)=
  \mathbf 1\{\text{accepted }b\}\,[\mu_t-F_t(b)]_+
  +
  \mathbf 1\{\text{rejected }b\}\,[F_t(b)-\mu_t]_+.
  \label{eq:audit-admission-regret}
\end{equation}
The first term measures avoidable pollution: accepting an incoming block that is worth
less than the cheapest legal displacement.  The second measures avoidable rejection:
rejecting a reachable block that is worth more than free space or the cheapest legal
displacement.  Descendants bypassed after a parent rejection are consequences of that
scored rejection and are not counted as additional admission decisions.

For an eviction event with selected victim \(e_t\), the local eviction-side surrogate
regret is
\begin{equation}
  R^E_t=
  F_t(e_t)-\min_{v\in\mathcal L(X_t)}F_t(v).
  \label{eq:audit-eviction-regret}
\end{equation}
This is value-weighted avoidable eviction.  It asks how much more future surrogate
value the chosen victim had than the cheapest legal victim in the same state.

Equations~\eqref{eq:audit-admission-regret}--\eqref{eq:audit-eviction-regret} decompose
a local same-state comparison.  They are not realized causal regret: changing one
admission or victim can change later reachability, pressure, legal victims, and even
which future requests become hits.

\subsection{Measured shadow-price tracking}

The local relaxation also yields a measurable control diagnostic.  Because one
physical cache block consumes one capacity slot, let \(w\) denote the configured
physical block size in token units and define the replay-oracle water level
\begin{equation}
  \lambda_t^\star=\frac{\mu_t}{w}.
  \label{eq:oracle-shadow-price}
\end{equation}
Candidate admission scores are dimensionless, so subtracting them directly from
\(\lambda_t^\star\) would be invalid.  For each workload-capacity-seed replay, fit a
single nonnegative scale with no intercept,
\begin{equation}
  \beta^\star=
  \arg\min_{\beta\ge0}
  \sum_t\left(
    \frac{F_t(b_t)-\mu_t}{w}-\beta s_t
  \right)^2,
  \label{eq:shadow-score-calibration}
\end{equation}
where \(s_t\) is the candidate's signed admission score.  The no-intercept constraint
preserves the actual policy boundary \(s_t=0\).  The candidate-implied water level and
tracking error are then
\begin{align}
  \widehat\lambda_t
  &=\frac{F_t(b_t)}{w}-\beta^\star s_t,\\
  e_t^\lambda
  &=\widehat\lambda_t-\lambda_t^\star
   =\frac{F_t(b_t)-\mu_t}{w}-\beta^\star s_t.
  \label{eq:shadow-tracking-error}
\end{align}
This is a quarantined hindsight diagnostic, not a deployable future-value estimate.
It measures whether the score behaves like an online surplus estimator after only a
unit calibration.  The validation tracking RMSE is retained as a MAP-Elites behavior
coordinate, but it is not added to the scalar selection objective.

The bandwidth prediction is tested groupwise.  A decision is ``fast'' when
\(|\lambda_t^\star-\lambda_{t-1}^\star|\) lies in the top quartile of positive changes
within its replay.  The strict claim requires every regretful moving group to place a
larger share of admission regret in this fast set than the set's decision share.
Across validation replays at 8, 16, 24, and 48 blocks for phase shifts, priority
bursts, tenant cycles, stochastic mixes, and hotset scans, the claim is falsified:
16 of 22 regretful moving groups satisfy it.  Aggregately, however, the fastest
3.5\% of decisions contain 12.5\% of admission regret, a \(3.96\times\) fast/slow
regret-density lift.  Tenant cycles satisfy the concentration test in 7 of 9 groups;
priority bursts in 4 of 6; stochastic mixes in 2 of 4.

The more important qualification is identifiability.  In 38 of 60 groups,
\(\lambda_t^\star\) never moves because a zero-future-value legal victim keeps
\(\mu_t=0\).  All 48-block groups and all phase-shift-prompt groups have this
degeneracy.  Thus pressure half-life is plausibly a loop-bandwidth parameter only in
capacity-constrained regimes where the local displacement price is informative.
The current evidence does not support treating threshold tracking as a universal
explanation of admission regret.

For group \(g\), define
\begin{align}
  A_g &= \sum_{t\in g}\sum_{b\in\mathcal R_t}R^A_t(b),\\
  E_g &= \sum_{t\in g}R^E_t,\\
  \bar A_g &= \frac{A_g}{\max(1,N^A_g)},
  \qquad
  \bar E_g = \frac{E_g}{\max(1,N^E_g)},
\end{align}
where \(\mathcal R_t\) is the set of explicitly scored reachable admission decisions,
\(N^A_g\) is the number of admission decisions, and \(N^E_g\) is the number of eviction
decisions.

The four dominance hypotheses are mathematically distinct:
\begin{align}
  H_{\mathrm{univ}} &: A_g>E_g\quad\forall g\in\mathcal G_+,\\
  H_{\mathrm{maj}} &: \frac{1}{|\mathcal G_+|}
    \sum_{g\in\mathcal G_+}\mathbf 1\{A_g>E_g\}>\frac12,\\
  H_{\mathrm{agg}} &: \sum_{g\in\mathcal G_+}A_g>
    \sum_{g\in\mathcal G_+}E_g,\\
  H_{\mathrm{rate}} &: \frac{1}{|\mathcal G_+|}
    \sum_{g\in\mathcal G_+}\mathbf 1\{\bar A_g>\bar E_g\}>\frac12,
\end{align}
where \(\mathcal G_+=\{g:A_g+E_g>0\}\).  The earlier universal wording corresponds to
\(H_{\mathrm{univ}}\), which the audit falsifies.

For \identifier{pressure\_aware\_incumbent}, the audit reports
\begin{align}
  |\mathcal G_+| &= 123,\\
  \sum_g A_g &= 181{,}327,\\
  \sum_g E_g &= 31{,}565,\\
  \frac{\sum_g A_g}{\sum_g(A_g+E_g)} &= 85.2\%,\\
  \sum_g\mathbf 1\{A_g>E_g\} &= 81,\\
  \sum_g\mathbf 1\{\bar A_g>\bar E_g\} &= 61.
\end{align}
Thus \(H_{\mathrm{agg}}\) is supported under this surrogate and \(H_{\mathrm{maj}}\) is
supported at \(81/123=65.9\%\).  But \(H_{\mathrm{univ}}\) is false because there are
\(42\) eviction-dominant regretful groups, and \(H_{\mathrm{rate}}\) is not supported
because \(61/123=49.6\%\).  The aggregate per-decision rate still favors admission,
\(8.56\) versus \(5.84\) future tokens per decision, which shows why totals and
per-decision dominance answer different questions.

The split-level summary is:
\begin{center}
\small
\begin{tabular}{lrrrrr}
\toprule
\textbf{Split} & \textbf{Groups} & \textbf{Adm.-dom.} & \textbf{Rate} &
\textbf{Admission} & \textbf{Eviction} \\
\midrule
hidden & 45 & 31 & 68.9\% & 30{,}182 & 18{,}255 \\
probe & 9 & 6 & 66.7\% & 49{,}490 & 822 \\
train & 24 & 12 & 50.0\% & 83{,}380 & 5{,}746 \\
validation & 45 & 32 & 71.1\% & 18{,}275 & 6{,}742 \\
\bottomrule
\end{tabular}
\end{center}

Eviction-dominant workload families include
\identifier{hidden/heavy\_tailed\_prefix\_lengths\_shifted},
\identifier{train/shared\_system\_prompt},
\identifier{validation/priority\_one\_off\_noise},
\identifier{train/rag\_template\_reuse}, and
\identifier{validation/multi\_tenant\_skew}.  These families should be treated as
positive evidence for eviction-specific follow-up, not as noise around an otherwise
uniform admission-first rule.

\subsection{Eviction opportunity, victim-spread, and why sophistication can matter}

The local eviction regret in Equation~\eqref{eq:audit-eviction-regret} is bounded by
the value spread of the legal victim set.  Define
\begin{equation}
  \delta^E_t=
  \max_{u,v\in\mathcal L(X_t)} |F_t(u)-F_t(v)|,
  \qquad
  n^E_t=|\mathcal L(X_t)|.
\end{equation}
Then \(R^E_t\le \delta^E_t\), and \(R^E_t=0\) whenever \(n^E_t\le 1\).  Eviction
sophistication can matter only on the opportunity set
\begin{equation}
  \mathcal O_E=\{t:n^E_t\ge 2\text{ and }\delta^E_t>0\}.
\end{equation}
The updated decision-level analysis shows that this opportunity set is large in the
current simulator: \(77.5\%\) of eviction decisions have multiple legal victims.

Let \(e^0_t\) be the incumbent victim and \(e^\phi_t\) the specialist victim in the
same state.  A same-state specialist correction occurs when
\begin{equation}
  F_t(e^\phi_t)<F_t(e^0_t),
\end{equation}
while an introduced mistake occurs when the inequality reverses.  The completed
specialist corrects 281 avoidable choices and introduces 150, for a net correction of
131.  In end-to-end replay it raises raw selection by \(1.101\), reduces validation
avoidable-eviction rate from \(0.1530\) to \(0.1350\), and reduces churn from
\(168.1\) to \(164.8\) per thousand requests.

This evidence changes the interpretation of eviction.  The correct statement is not
that eviction has no room to improve.  The correct statement is that the currently
found richer eviction improvements sit on an unfavorable complexity-generalization
frontier.  The full function-only specialist has real victim-ranking signal but
composes to \(1{,}023\) nodes.  Guarded age plus descendant protection captures
\(98.0\%\) of the full specialist's raw gain, but still composes to 787 nodes.  A
single deployable distillation--raising descendant protection from \(0.2\) to
\(0.6\)--keeps complexity at 648 and gains \(0.168\) raw selection, but loses
\(0.006\) hidden score.  Hence eviction is behaviorally meaningful, while promotable
eviction sophistication remains unproven.

\subsection{When admission dominates and when eviction dominates}

The model predicts different dominant error modes from two local quantities:
\begin{align}
  \gamma^A_t(b)&=|F_t(b)-\mu_t|,
  \label{eq:admission-margin-regime}\\
  \delta^E_t&=\max_{u,v\in\mathcal L(X_t)}|F_t(u)-F_t(v)|.
  \label{eq:eviction-spread-regime}
\end{align}
The admission margin \(\gamma^A_t(b)\) measures how costly it is to classify an
incoming block on the wrong side of the displacement threshold.  The eviction spread
\(\delta^E_t\) measures how costly it is to choose the wrong victim among legal leaves.

Admission-side control should dominate when:
\begin{itemize}
  \item many reachable blocks have large threshold margins and are misclassified;
  \item low-value admitted blocks persist long enough to cause later evictions or
  underfill;
  \item pressure shocks raise \(\mu_t\) or \(\lambda_t\), making old thresholds too
  permissive;
  \item parent admission opens a deep suffix of low-value descendants.
\end{itemize}

Eviction-side control should dominate when:
\begin{itemize}
  \item \(n^E_t\) is large and \(\delta^E_t\) is large;
  \item legal leaves have sharply different remaining reuse despite similar recency;
  \item a valuable shared-root or template branch appears as an inactive leaf;
  \item priority or one-off traffic pollutes frequency and age signals;
  \item heavy-tailed prefix lengths create high-value deep leaves beside disposable
  shallow leaves.
\end{itemize}

This explains the audit.  Aggregate admission regret can be large because every missed
reachable block creates a threshold decision.  Eviction can still dominate specific
families because a smaller number of wrong victim choices can destroy high-value
future reuse.

\subsection{A regime-gated hybrid policy model}

The crossed replay supports a sparse hybrid policy family rather than a larger
unconditional ranker.  Let \(h_t\in\{0,1\}\) be an online regime gate estimated from
candidate-visible
features such as legal-victim count, age dispersion, descendant distribution, observed
frequency dispersion, pressure, miss rate, and priority-backed reuse.  A composable
eviction score can be written as
\begin{equation}
  \varepsilon_t(b)=
  (1-h_t)\varepsilon^{\mathrm{compact}}_t(b)
  +h_t\varepsilon^{\mathrm{specialist}}_t(b).
  \label{eq:regime-gated-eviction}
\end{equation}
The gate should activate only when the state resembles the eviction-opportunity set:
multiple legal victims, large estimated value spread, and a workload signature known to
produce eviction-dominant regret.  Otherwise the compact frequency-aware ranker remains
preferred because it is cheaper, more stable, and already close to the best deployable
baseline.

The same idea applies to admission.  The pressure-aware incumbent estimates a shadow
price \(\lambda_t\) from recent admission pressure and miss rate, then raises the
threshold for deep low-evidence blocks.  A stronger policy should estimate not only a
single pressure scalar but also the uncertainty around the incoming block value.  In
states with high admission uncertainty and low victim spread, selective bypass is the
right conservative action; in states with high victim spread, the policy should spend
more complexity on victim ranking.

\subsection{Causal component experiments}

The local audit is not enough to establish causal component value.  We therefore ran
the crossed replay using incumbent and oracle components.  Let
\(Y_{AE}(g)\) be the group score when admission component \(A\in\{I,O\}\) and eviction
component \(E\in\{I,O\}\) are selected from the incumbent \(I\) or a legal future-aware
oracle \(O\).  The paired effects are
\begin{align}
  \tau_A(g)&=Y_{OI}(g)-Y_{II}(g),\\
  \tau_E(g)&=Y_{IO}(g)-Y_{II}(g),\\
  \tau_{AE}(g)&=Y_{OO}(g)-Y_{OI}(g)-Y_{IO}(g)+Y_{II}(g).
\end{align}
Table~\ref{tab:causal-component-factorial} reports the completed full-panel result.
The primary outcome is the evaluator's behavioral group score with the future-aware
avoidable-eviction penalty and common complete-policy complexity charge omitted.  This
prevents the causal outcome from directly rewarding agreement with the audit surrogate.
The replay covers 162
workload--capacity--seed groups with 96 requests per workload, capacities 24 and 48,
and three seeds.  The same conclusions hold directionally for token hit rate and
prefill tokens saved in the machine-readable artifact.

\begin{table}[H]
\centering
\small
\caption{Paired causal component effects.  ``Eviction after \(O_A\)'' is
\(Y_{OO}-Y_{OI}=\tau_E+\tau_{AE}\).}
\label{tab:causal-component-factorial}
\begin{tabular}{lrrrrr}
\toprule
Group set & \(n\) & \(\sum\tau_A\) & \(\sum\tau_E\) &
\(\sum\tau_{AE}\) & Eviction after \(O_A\) \\
\midrule
Full panel & 162 & 1704.963 & 518.431 & -258.918 & 259.513 \\
Audit-labeled eviction-dominant & 29 & 530.681 & 419.871 & -340.599 & 79.271 \\
\bottomrule
\end{tabular}
\end{table}

Admission therefore has \(3.29\times\) the aggregate eviction main effect and is
larger in \(136/162\) groups.  The interaction directly rejects a literal reading of
the surrogate counterexamples: among the 29 groups labeled eviction-dominant by the
same-state audit, admission is causally larger in 23, oracle admission removes
\(81.1\%\) of aggregate eviction value, and 18 retain no positive eviction effect
after admission is repaired.  Across the full panel, however, \(37/162\) groups retain
positive residual eviction value and \(50.1\%\) of the aggregate eviction main effect
remains.  The causal conclusion is consequently asymmetric: optimize admission first,
then gate richer eviction logic only into the minority regimes where its value
survives the admission intervention.  Detailed workload effects and all secondary
outcomes are preserved in \filepath{docs/research_log.tex}; exact paired results remain
in \filepath{artifacts/prefix_kv_cache_causal_component_factorial.json}.

A second, cheaper experiment is a factorial matrix of deployable components: several
admission rules crossed with several eviction rules, all replayed on the same traces.
The current result already points in this direction: pure LRU eviction leaves measurable
mean token-hit rate unused across tested non-oracle admission rules, while LFU and the
incumbent compact value-aware eviction ranker are close.  This supports an asymmetric
engineering default--selective, regime-aware admission with simple frequency-aware
eviction--but not a proof that eviction is solved.

\subsection{Predictions and next audits}

The regime-conditional model makes testable predictions:

\begin{enumerate}
  \item \textbf{Victim-spread prediction.}  Eviction-dominant groups should have larger
  \(\delta^E_t\), larger legal-victim counts, and worse chosen-victim value ranks than
  admission-dominant groups.
  \item \textbf{Complexity-frontier prediction.}  Specialist variants should improve
  same-state victim rank before they improve charged complete-policy score.  Promotion
  requires distilling that ranking signal below the 650-node gate without hidden loss.
  \item \textbf{Admission-threshold prediction.}  Admission-dominant groups should show
  many large-margin threshold mistakes \(\gamma^A_t\), especially under pressure shifts,
  agent traces, session continuations, phase shifts, and rolling-template workloads.
  \item \textbf{Interaction prediction.}  Oracle admission should reduce the number and
  spread of later eviction opportunities by preventing pollution; oracle eviction may
  matter most when incumbent admission has already created a mixed-value legal-victim
  set.
  \item \textbf{Decode-pressure prediction.}  When active decode KV consumes shared
  capacity, prefix-only admission thresholds should be too permissive unless they treat
  decode occupancy as part of the shadow price.
\end{enumerate}

The immediate logging additions should be: chosen-victim value rank, legal-victim value
spread, legal-victim count, whether the chosen victim lies on a high-reuse branch,
admission displacement threshold \(\mu_t\), and scored-block margin \(F_t(b)-\mu_t\).
Those fields would let the report state not just that admission or eviction dominated a
group, but why the corresponding local decisions were hard.

\begin{thebibliography}{99}
\addcontentsline{toc}{section}{References}

\bibitem{pagedattention}
W. Kwon et al.
\newblock ``Efficient Memory Management for Large Language Model Serving with
PagedAttention.''
\newblock ACM SOSP, 2023.
\newblock \url{https://arxiv.org/abs/2309.06180}.

\bibitem{vllm-apc}
vLLM Project.
\newblock ``Automatic Prefix Caching: Design.''
\newblock Accessed June 5, 2026.
\newblock \url{https://docs.vllm.ai/en/stable/design/prefix_caching/}.

\bibitem{sglang}
L. Zheng et al.
\newblock ``Efficiently Programming Large Language Models using SGLang.''
\newblock arXiv preprint, version 1, 2023.
\newblock \url{https://arxiv.org/html/2312.07104v1}.

\bibitem{saecache}
S. Fang et al.
\newblock ``Not All Tokens Are Worth Caching: Learning Semantic-Aware Eviction for LLM
Prefix Caches.''
\newblock arXiv preprint, May 2026.
\newblock \url{https://arxiv.org/abs/2605.18825}.

\bibitem{kvflow}
Z. Pan et al.
\newblock ``KVFlow: Efficient Prefix Caching for Accelerating LLM-Based Multi-Agent
Workflows.''
\newblock arXiv preprint, 2025.
\newblock \url{https://arxiv.org/abs/2507.07400}.

\bibitem{infercept}
R. Abhyankar et al.
\newblock ``INFERCEPT: Efficient Intercept Support for Augmented LLM Inference.''
\newblock ICML, 2024.
\newblock \url{https://arxiv.org/abs/2402.01869}.

\bibitem{preble}
V. Srivatsa et al.
\newblock ``Preble: Efficient Distributed Prompt Scheduling for LLM Serving.''
\newblock arXiv preprint, 2024.
\newblock \url{https://arxiv.org/abs/2407.00023}.

\bibitem{mooncake}
R. Qin et al.
\newblock ``Mooncake: A KVCache-centric Disaggregated Architecture for LLM Serving.''
\newblock arXiv preprint, 2024.
\newblock \url{https://arxiv.org/abs/2407.00079}.

\bibitem{memserve}
C. Hu et al.
\newblock ``MemServe: Context Caching for Disaggregated LLM Serving with Elastic Memory
Pool.''
\newblock arXiv preprint, 2024.
\newblock \url{https://arxiv.org/abs/2406.17565}.

\bibitem{lmcache}
Y. Liu et al.
\newblock ``LMCache: An Efficient KV Cache Layer for Enterprise-Scale LLM Inference.''
\newblock arXiv preprint, 2025.
\newblock \url{https://arxiv.org/abs/2510.09665}.

\bibitem{tutti}
S. Qiu et al.
\newblock ``Tutti: Making SSD-Backed KV Cache Practical for Long-Context LLM Serving.''
\newblock arXiv preprint, May 2026.
\newblock \url{https://arxiv.org/abs/2605.03375}.

\bibitem{promptcache}
I. Gim et al.
\newblock ``Prompt Cache: Modular Attention Reuse for Low-Latency Inference.''
\newblock arXiv preprint, 2023.
\newblock \url{https://arxiv.org/abs/2311.04934}.

\bibitem{ragcache}
C. Jin et al.
\newblock ``RAGCache: Efficient Knowledge Caching for Retrieval-Augmented Generation.''
\newblock arXiv preprint, 2024.
\newblock \url{https://arxiv.org/abs/2404.12457}.

\bibitem{cacheblend}
J. Yao et al.
\newblock ``CacheBlend: Fast Large Language Model Serving for RAG with Cached Knowledge
Fusion.''
\newblock arXiv preprint, 2024.
\newblock \url{https://arxiv.org/abs/2405.16444}.

\bibitem{h2o}
Z. Zhang et al.
\newblock ``H2O: Heavy-Hitter Oracle for Efficient Generative Inference of Large
Language Models.''
\newblock NeurIPS, 2023.
\newblock \url{https://arxiv.org/abs/2306.14048}.

\bibitem{kivi}
Z. Liu et al.
\newblock ``KIVI: A Tuning-Free Asymmetric 2bit Quantization for KV Cache.''
\newblock ICML, 2024.
\newblock \url{https://arxiv.org/abs/2402.02750}.

\bibitem{snapkv}
Y. Li et al.
\newblock ``SnapKV: LLM Knows What You Are Looking for Before Generation.''
\newblock NeurIPS, 2024.
\newblock \url{https://arxiv.org/abs/2404.14469}.

\bibitem{quest}
J. Tang et al.
\newblock ``Quest: Query-Aware Sparsity for Efficient Long-Context LLM Inference.''
\newblock ICML, 2024.
\newblock \url{https://arxiv.org/abs/2406.10774}.

\bibitem{laprox}
T. Mai and J.-Y. Kim.
\newblock ``Reformulating KV Cache Eviction Problem for Long-Context LLM Inference.''
\newblock arXiv preprint, May 2026.
\newblock \url{https://arxiv.org/abs/2605.07234}.

\bibitem{vericache}
J. Yao et al.
\newblock ``VeriCache: Turning Lossy KV Cache into Lossless LLM Inference.''
\newblock arXiv preprint, May 2026.
\newblock \url{https://arxiv.org/abs/2605.17613}.

\end{thebibliography}

\end{document}
